---
title: "Data Science Sampler: Beyond Static Plots"
author: "Casey Schine"
format:
  html:
    toc: true
    code-fold: show
    code-tools: true
    embed-resources: true
execute:
  message: false
  warning: false
editor_options: 
  chunk_output_type: console
---

## Overview

In this class, we've focused on building solid data analysis skills.

Today, we’re going to look at some *optional tools* that go beyond that—things that make your work:

- interactive  
- public-facing  
- more engaging  

You will **vote on which of these we explore further**.

---

## Setup

Here is where you may need to install some new packages:

```{r warning=FALSE, message=FALSE}
# Uncomment if needed
# install.packages(c("tidyverse", "leaflet", "plotly", "gganimate",
#                    "gapminder", "palmerpenguins"))

library(tidyverse)
library(leaflet) #new
library(plotly) #new
library(gganimate) #new
library(gapminder)
library(palmerpenguins)


```

## Leaflet Interactive Maps

Let's start with a basic map of earthquakes near New Zealand. We can set the earthquakes to size by magnitude. I messed around with the math a little to get a size ratio that I liked.

```{r}
data(quakes)

quakes |>
  slice_sample(n = 300) |>
  leaflet() |>
  addProviderTiles("CartoDB.Positron") |>
  addCircleMarkers(
    lng = ~long,
    lat = ~lat,
    # set circle marker to size by magnitude
    radius = ~mag^2/3,
    # make popup labels
    popup = ~paste0(
      "<strong>Magnitude:</strong> ", mag,
      "<br><strong>Depth:</strong> ", depth, " km"
    ),
    stroke = FALSE,
    fillOpacity = 0.6
  )
```
 
Color the points by magnitude and add a legend: 

```{r}
pal <- colorNumeric(
  palette = "viridis",
  domain = quakes$mag
)

leaflet(quakes) |>
  addProviderTiles("CartoDB.Positron") |>
  addCircleMarkers(
    lng = ~long,
    lat = ~lat,
    radius = ~mag^2/3,
    color = ~pal(mag),
    stroke = FALSE,
    fillOpacity = 0.7
  ) |>
  addLegend(
    "bottomright",
    pal = pal,
    values = ~mag,
    title = "Magnitude"
  )

```


Now let's start playing around with the basemaps. You can find all of the possibilities on this website: 

https://leaflet-extras.github.io/leaflet-providers/preview/

```{r}

library(leaflet)
library(dplyr)

data(quakes)

small <- quakes |> filter(mag < 5)
large <- quakes |> filter(mag >= 5)

leaflet() |>
  # Basemaps
  addProviderTiles("CartoDB.Positron", group = "Light map") |>
  addProviderTiles("Esri.WorldImagery", group = "Satellite") |>
  addProviderTiles("Stadia.StamenWatercolor", group = "Artsy") |>

  # Small earthquakes
  addCircleMarkers(
    data = small,
    lng = ~long, lat = ~lat,
    radius = 3,
    color = "blue",
    group = "Smaller quakes",
    stroke = FALSE,
    fillOpacity = 0.5
  ) |>

  # Large earthquakes
  addCircleMarkers(
    data = large,
    lng = ~long, lat = ~lat,
    radius = ~mag * 2,
    color = "red",
    group = "Larger quakes",
    stroke = TRUE,
    weight = 1
  ) |>

  # Layer control
  addLayersControl(
    baseGroups = c("Light map", "Satellite", "Artsy"),
    overlayGroups = c("Smaller quakes", "Larger quakes"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

Now with the EJI data. I strongly recommend that you don't try to run this one. 

```{r}
library(sf)
eji_data <- st_read("~/Dropbox/!Middlebury/Teaching/STAT 118/S2026/S2026_R_Proj/midterm/eji_sf.gpkg")

eji_cali <- eji_data |>
  filter(STATE_NAME=="California") |>
  filter(!is.na(E_LIMENG), E_LIMENG <100)

pal <- colorNumeric(
  palette = "viridis",
  domain = eji_cali$E_LIMENG,
  na.color = "transparent"
)


eji_cali |>
leaflet() |>
  addProviderTiles("CartoDB.Positron") |>
  addPolygons(
    fillColor = ~pal(E_LIMENG),
    fillOpacity = 0.7,
    color = "white",
    weight = 0.5,
    popup = ~paste0(
      "<strong>County:</strong> ", COUNTY,
      "<br><strong>Tract:</strong> ", GEOID,
      "<br><strong>Poverty %:</strong> ", E_LIMENG
    )
  ) |>
  addLegend(
    "bottomright",
    pal = pal,
    values = ~E_LIMENG,
    title = "Unemployed",
    opacity = 0.7
  )
```


## Plotly Interactive Plots

Plotly doesn’t change your analysis at all — it just changes how your results exist. It turns a static figure into something you could actually put on a website or share with someone to explore.

Let's start simple. You make a plot in ggplot and then you feed it into plotly and it's like magic.

```{r}
p <- ggplot(mpg, aes(x = displ, y = hwy, color = class,
                     text = paste("Model:", model))) +
  geom_point(size = 2, alpha = 0.7) +
  theme_minimal()

ggplotly(p, tooltip = "text")

```

Let's add more information to the pop-up:

```{r}
p <- ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point(
    aes(text = paste(
      "Model:", model,
      "<br>Class:", class,
      "<br>Displacement:", displ,
      "<br>Highway MPG:", hwy
    )),
    size = 2,
    alpha = 0.7
  ) +
  theme_minimal()

ggplotly(p, tooltip = "text")
```

Let's try it with the gapminder data. 

```{r}
library(gapminder)

p <-  gapminder |> 
  filter(year == 2007) |>
  ggplot(aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point(alpha = 0.7) +
  labs(color="Continent") +
  scale_x_log10() +
  scale_size(range = c(2, 10)) +
  ggtitle("Year: 2007") +
  theme_minimal()

ggplotly(p)
```

With facets:

```{r}
p <- ggplot(mpg, aes(displ, hwy, color = class)) +
  geom_point(alpha = 0.7) +
  facet_wrap(~drv) +
  theme_minimal()

ggplotly(p)
```

With a boxplot:

```{r}
p <- ggplot(mpg, aes(class, hwy, fill = class)) +
  geom_boxplot() +
  theme_minimal()

ggplotly(p)
```

This is my first time doing this with a boxplot, and now this is the only way I'm making a boxplot ever again!

Let's do it with a line!

```{r}
economics <- ggplot2::economics

p <- ggplot(economics, aes(date, unemploy)) +
  geom_line(color = "steelblue") +
  theme_minimal()

ggplotly(p)
```

WITH A DENSITY PLOT!

```{r}
p <- ggplot(diamonds, aes(x = price, fill = cut)) +
  geom_density(alpha = 0.5) +
  theme_minimal()

ggplotly(p)
```

This is AMAZING! Why am I only just learning this????


## Animation

This animation shows the gapminder dataset through time, with GDP per capita on the x-axis and Life expectancy on the y-axis. The points are colored by continent and sized by population (on a log scale).

```{r eval=FALSE}

p <- ggplot(
  gapminder,
  aes(
    x = gdpPercap,
    y = lifeExp,
    group = country,
    color = continent,
    size = pop
  )
) +
  geom_point(alpha = 0.7) +
  scale_x_log10() +
  scale_color_brewer(palette = "Set1") +
  scale_size_continuous(range = c(1, 10)) +
  theme_minimal() +
  labs(
    title = "Year: {frame_time}",
    x = "GDP per capita",
    y = "Life expectancy",
    color = "Continent",
    size = "Population"
  ) +
  transition_time(year)

animate(
  p,
  nframes = 60,
  fps = 4,
  width = 600,
  height = 400,
  renderer = gifski_renderer(),
  device = "ragg_png"
)
```

You may be able to tell by how many plots I made with the others and how few animations I have given you, that my heart isn't really in animations. This single animation took about 50% of the time I spent making this. They can be a real pain. A good animation is absolute gold, but it is very hard to make a good animation. 

## Spatial Data - but not people data

```{r}
library(terra)
library(tidyterra)
library(cmocean)
library(ggthemes)
library(wesanderson)

volcano_df <- as.data.frame(volcano) |>
  mutate(y = row_number()) |>
  pivot_longer(-y, names_to = "x", values_to = "elevation") |>
  mutate(x = as.numeric(gsub("V", "", x)))

ggplot(volcano_df, aes(x, y, fill = elevation)) +
  geom_raster() + #not the way I choose to do this anymore, but workable for smaller data
  scale_fill_cmocean(name="thermal") +
  coord_equal() +
  scale_y_reverse() +
  theme_minimal() +
  labs(title = "Topographic-style raster map")

```

Here is some code from my Data Science Across Disciplines class this last January looking at instrumented seal data from the Southern Ocean. 

First, some data prep. You are not expected to understand this part. 

```{r data_prep, cache=TRUE, message=FALSE}

seal_data <- read_csv("~/Dropbox/!middlebury/Teaching/DSAD 1230/DSAD_2026/R_code/MEOP_data_analysis/data/regional_data/par_seal_data.csv")
seal_data_surface <- seal_data |>
  filter(level==1)

NPP_stack <- rast("~/Dropbox/!Research/Projects/PacAntRidge/data/NPP_raster_stack.nc")

invisible(capture.output({ #to suppress the #$^*@$ progress bar
NPP_stack[NPP_stack >100] = 100 #set max value
NPP_stack[NPP_stack <0] = NA #transform placeholder NA value (-99) to actual NA
}))
raster_dir <- "/Users/caseysmith/Dropbox/!Middlebury/Teaching/DSAD 1230/DSAD_2025/R_code/Argo_vis_shiny_app"

PacSec_bathy <- rast(paste0(raster_dir, "/data/raster/PacSec_bathy.tif"))


#adjust seal data
seal_data_surface <- seal_data_surface |>
  mutate(lon_anti=case_when(lon<0 ~ lon+360,
                            lon>=0 ~lon))

seal_data_mod <- seal_data |>
  mutate(lon_anti=case_when(lon<0 ~ lon+360,
                            lon>=0 ~lon))

world <- map_data("world")
```

First a map with "raster" data. 

```{r map_bathy_with_seals, message=FALSE, fig.height=4, fig.width=10}
ggplot() +
  geom_spatraster(data=PacSec_bathy) +
  scale_fill_cmocean(name="gray",
                     limits=c(-6000,0),
                     breaks=c(-6000,-5000,-4000,-3000,-2000,-1000,0),
                     labels=c("-6","-5","-4", "-3","-2","-1", "0")) +
  geom_polygon(data = world, 
               aes(x = long, y = lat, group = group),
               fill="white", color="white") +
   labs(fill="Depth (km)",
       x="Longitude", y="Latitude") +
  coord_sf(expand=FALSE, xlim = c(140,220), ylim=c(-70,-55) ) +
  ggtitle("Zoomed in on AAR/PAR, Bathymetry") +
  theme_tufte()

ggplot() +
  geom_spatraster(data=PacSec_bathy) +
  scale_fill_cmocean(name="gray",
                     limits=c(-6000,0),
                     breaks=c(-6000,-5000,-4000,-3000,-2000,-1000,0),
                     labels=c("-6","-5","-4", "-3","-2","-1", "0")) +
  geom_polygon(data = world, 
               aes(x = long, y = lat, group = group),
               fill="white", color="white") +
  geom_point(data=seal_data_surface, 
             aes(x=lon_anti, y=lat, color=TEMP_ADJUSTED), 
             alpha=0.3, size=2) +
  scale_color_cmocean(name="thermal") +
  labs(fill="Depth (km)", color="Temp",
       x="Longitude", y="Latitude") +
  coord_sf(expand=FALSE, xlim = c(140,220), ylim=c(-70,-55) ) +
  ggtitle("Zoomed in on PAR, Bathymetry and Temperature") +
  theme_tufte()
```

```{r NPP_layer_plot, message=FALSE, warning=FALSE, fig.height=4, fig.width=9}
#my favorite palette from the wesanderson package
pal <- wes_palette("Zissou1", 100, type = "continuous")
ggplot() +
  geom_spatraster(data=NPP_stack[[15]]) + #index to show layer 15 only
  scale_fill_gradientn(colors=pal, limits=c(0,100)) +
  geom_spatraster_contour(data=PacSec_bathy, breaks=c(0,-3000)) + #contour to show margins of mid ocean ridge
  labs(fill="NPP",
       x = "Longitude",
       y = "Latitude") +
  xlim(130, 230) +
  ylim(-70, -55) +
  coord_sf(expand=FALSE) +
  theme_tufte() +
  ggtitle("NPP 2011-2012")
```

```{r calc_diff_layer, message=FALSE}
#calculate the difference between the two years, by subtracting one from the other
diff_layer <- NPP_stack[[14]]-NPP_stack[[15]]

diff_layer[diff_layer > 75] = 75 #set max value
diff_layer[diff_layer < (-75)] = -75 #set min value
```

```{r NPP_diff_plot, message=FALSE, fig.height=4, fig.width=9}

ggplot() +
  geom_spatraster(data=diff_layer) + 
  scale_fill_cmocean(name="balance", limits=c(-75,75)) + #change to diverging palette
  geom_spatraster_contour(data=PacSec_bathy, breaks=c(0,-3000)) +
  labs(fill="NPP",
       x = "Longitude",
       y = "Latitude") +
  xlim(130, 230) +
  ylim(-70, -55) +
  coord_sf(expand=FALSE) + #set this argument so that it doesn't add a border between the map and the axes
  theme_tufte() +
  ggtitle("NPP difference 2010-2011 minus 2011-2012")
```

## Schiny

This is just some example code. We're not going to run this. I'm going to show you an example from students I had during J-term. 

```{r eval=FALSE, echo=TRUE}
library(shiny)
library(ggplot2)
library(dplyr)
library(palmerpenguins)

ui <- fluidPage(
  selectInput("species", "Species:",
              choices = c("All", unique(na.omit(penguins$species)))),
  plotOutput("plot")
)

server <- function(input, output) {

  filtered <- reactive({
    if (input$species == "All") penguins
    else filter(penguins, species == input$species)
  })

  output$plot <- renderPlot({
    ggplot(filtered(),
           aes(bill_length_mm, flipper_length_mm, color = species)) +
      geom_point() +
      theme_minimal()
  })
}

shinyApp(ui, server)
```

## Github Website

This is a fun one that I would love for you to choose because I really want to learn it. This would also be a very cool thing to be able to put on a resume and showcase all of your cool plots that you've made!



