---
title: '`forcats`: working with categorical data'
author: 'Casey Schine adapted from Emily Malcolm-White'
format:
  html: 
    toc: TRUE
    code-overflow: wrap
    embed-resources: true
    code-tools:
      source: true
      toggle: false
      caption: none
execute: 
  message: FALSE
  warning: FALSE
---

The R package `forcats` is designed to make working with categorical variables easier and more efficient. It provides a set of functions that allow you to manipulate and analyze categorical data with ease. In this lesson, we'll cover the basics of the `forcats` package and some of its most useful functions.

## Categorical Variables and Factors

Categorical variables contain values that represent groups or categories. Some categories have no inherent order (nominal), while others have a meaningful order (ordinal).

Examples of categorical data include:

-   Colors (red, blue, green, etc.)
-   Types of vehicles (sedan, SUV, truck)
-   Educational degrees (high school, college, graduate school)

In R, a **factor** is a way of storing categorical data that keeps track of both the categories and their order. These categories are called levels.

## `mpg` Data

We will play with different functions in the `forcats` packages using the `mpg` dataset from earlier in the semester.

```{r}
library(forcats)
library(tidyverse)
data("mpg")
library(ggthemes)
```

Let's take a look at a side-by-side boxplot:

```{r}

```

## Reordering Factor Levels

One of the most useful functions is `fct_relevel()`, which allows you to reorder the levels of a factor. This can be useful when you want to change the default ordering of the levels.

Is `class` a factor?

```{r}
# look at the mpg data and the column classes

# use is.factor() to inquire if class is a factor variable
# returns TRUE or FALSE

# with dplyr syntax (pipes)
# with pull()


```

Let's make it a factor!

```{r}
# use mutate to write over class with new factor variable called class
# use as.factor()


```

Let's check the levels and their current ordering!

```{r}

# use levels() to get factor levels
# levels() requires vector so must use pull() rather than select()



```

### Reorder the levels with `fct_relevel()`

```{r}
# what do we think would be a better order
# reorder using fct_relevel() 
# use mutate to overwrite the existing class column



# check the levels

```

Let's recreate our side-by-side boxplot now:

```{r}

```

### Reorder levels with `fct_reorder()`

Rather than reordering them manually by typing the order, you could also re-level by some numeric criteria using `fct_reorder()`. For example:

```{r}
#?fct_reorder

# reorder class using median hwy mpg 
# as fct_reorder(class, hwy, median)
# fct_reorder(factor, values, function)



# take a look at the order again


# remake the boxplot again with the new order


# what other functions can we use to reorder???

```

## Renaming Factor levels with `fct_recode`

Sometimes you might not like the way the levels are named.

```{r}
# use fct_recode to change the name of a factor level in class
# fct_recode("two-seater" = "2seater")
## NEW NAME = OLD NAME



# take another look at the factor


```

## Factor Collapsing with `fct_collapse()`

Let's say we wanted to create only two categories -- cars and larger vehicles.

```{r}

# use fct_collapse to collapse class into cars and big factors
# mutate on new variable called class_two
# fct_collapse(class, cars = c("compact", "subcompact", "midsize", "two-seater"), big = c("pickup", "suv", "minivan"))


# take a look at the new variable factor levels

```

## Lumping into an other category

-   `fct_lump_min()`: lumps levels that appear fewer than min times.

-   `fct_lump_prop()`: lumps levels that appear in fewer than (or equal to) prop \* n times.

-   `fct_lump_n()` lumps all levels except for the n most frequent (or least frequent if n \< 0)

```{r}
# count manufacturers in dataset


```


Let's say we wanted only the manufacturers with at least 15 cars produced. Everything else we want to just be other:

```{r}
# ?fct_lump_min

# make manufacturer lump variable using fct_lump_min(15)
# lumps anything that does not have a minimum number of entries


# look at the result with count()


# now with fct_lump_n() for the top 5
# lumps everything that isn't the 5 highest manufacturers



# look at the result with count()




```

## Kable Tables

Create a table using kableExtra:

```{r}
library(kableExtra)

# make table for manufacturer before and after lumping 
# using kbl() |> kable_styling()



# using lumped manufacturers

```

If we want to change the order of items that appear in our table we can use some of the tricks that we just learned for reordering factors! 


We can also use Kable tables to display summary data.

```{r}

# we can use group_by() and summarize() to get statistics for each group
# calculate median hwy miles per gallon



# we can also use the fct_reorder() function to reorder the factor by a calculated value (as we did above)
# class = fct_reorder(class, hwy, .fun = median)
# calculate median hwy and add count using the n() function in summarize



# add caption for the table using the caption agument in kbl()

```

We can also control aspects of the table and how it looks using arguments in the `kbl()` and `kable_styling()` functions. 

```{r}

# how can we control aspects of the table 
#?kbl to look at some of the options

# can set the number of digits using the digits=x argument in kbl()

# also col.names argument to change the column names for the output table




```


## Heatmaps

A heatmap is a great way to summarize information across categories. Heatmaps use **color** to represent a numeric value for each combination of two categorical variables.

Let's start by looking at two categorical variables: vehicle `class` and drive type (`drv`).

```{r}
# count the number of observations in each combination
# of class and drive type


```

We can display these counts as a heatmap using `geom_tile()`:

```{r}
# count(class, drv) and set x to drv, y to class, and fill to n


```

Each tile represents one combination of vehicle class and drive type. The color of the tile represents the number of observations in that combination.


### Heatmaps of summary values

The color does not have to represent a count. We can calculate a summary statistic for each combination of categories and use that value instead.

```{r}
# calculate mean highway MPG for each combination
# of class and drive type


```

Now we can map `mean_hwy` to `fill` instead:

```{r}

```

Heatmaps are `ggplot` figures, so we can customize them using the same tools we have already learned. We can change axis and legend labels with `labs()`, choose a theme, and control the fill colors using a `scale_fill_*()` function.

## Explore the Youth Risk Behavior Survey

### 1. Understand the dataset
Look through the YRBS documentation and briefly describe what the dataset contains and what one row represents.

### 2. Choose a variable 
Find one variable that interests you. What does it measure, and what are its possible values? 


### 3. Summarize it
Use `count()` or `group_by()` + `summarize()` to explore your variable.

### 4. Make a figure 
Choose a second variable and create a visualization that helps you explore how the two variables are related.

### 5. Ask a question
Based on what you found, write one question about the YRBS data that you would be interested in investigating further. 
