---
title: 'Customizing Plots'
subtitle: '`scales`, labels, `facet_wrap()`'
author: 'Casey Schine adapted from Emily Malcolm-White'
format:
  html: 
    toc: TRUE
    code-overflow: wrap
    embed-resources: true
    code-tools:
      source: true
      toggle: false
      caption: none
    code-annotations: hover
execute: 
  message: FALSE
  warning: FALSE
editor_options: 
  chunk_output_type: console
---


```{r}
#| warning: FALSE
#| message: FALSE
library(tidyverse)
library(ggthemes)
```

```{r}
#Import the can_lang dataset 
can_lang <- read.csv("https://raw.githubusercontent.com/ttimbers/canlang/master/inst/extdata/can_lang.csv")
```

## A starting graph: scatterplot of `can_lang`

```{r}

# make base plot (scatter) with x as most_at_home and y as mother_tongue

can_lang |>
  ggplot(aes(x=most_at_home, y= mother_tongue)) +
  geom_point()




# Assign the plot to an object so it can be recalled
# Use these labels for the axis labels. 
  #xlab("Language spoken most at home \n (number of Canadian residents)") +
  #ylab("Mother tongue \n (number of Canadian residents)") +

can_lang_baseplot <- can_lang |>
  ggplot(aes(x=most_at_home, y= mother_tongue)) +
  geom_point()

can_lang_baseplot


# recall the baseplot and change the theme
 
can_lang_baseplot +
  theme_tufte()

# set the theme on the baseplot 
can_lang_baseplot <- can_lang |>
  ggplot(aes(x=most_at_home, y= mother_tongue)) +
  geom_point() +
  xlab("Language spoken most at home \n (number of Canadian residents)") +
  ylab("Mother tongue \n (number of Canadian residents)") +
  theme_tufte()
  

```

What do you notice about this plot?

::: {.callout-note icon="false"}
Note that I assigned this plot to an object and then called the object by name. This makes it possible to build on the plot if you want.
:::

## Axis display format: `scales` package

```{r}
 # Install the package if needed
library(scales)

```

We want to customize how the continuous x and y axes look, so we need to use the argument `labels=label_comma()` inside a `scale_*_continuous()` layer:

```{r}
# same base plot, change to scale_*_continuous and use the labels argument to control the appearance of the axes
# set labels=label_comma()

can_lang_baseplot +
  scale_x_continuous(labels=label_comma()) +
  scale_y_continuous(labels=label_comma())



```

::: callout-note
# What other formats are available in the `scales` package?

When passing a formatting function inside `scale_*_continuous(labels = ...)` you have options!

| Function | Use Case | Example Input | Example Output |
|------------------|------------------|------------------|------------------|
| `label_comma()` | Formats numbers with commas | `1234567` | `"1,234,567"` |
| `label_dollar()` | Formats numbers as dollar currency | `99.99` | `"$99.99"` |
| `label_dollar(prefix = "€")` | Formats numbers as euro currency | `99.99` | `"99.99€"` |
| `label_percent()` | Converts decimals to percent | `0.25` | `"25%"` |
| `label_pvalue()` | Formats p-values | `0.00005` | `"<0.0001"` |
:::


### Logarithmic Axes Transformations

::: callout-note
# Applying a Log Transformation

When you apply a log transformation to an axis (or both axes) in a plot, you convert values using a logarithmic scale instead of a linear scale. This means:

-   Instead of evenly spaced values (1, 2, 3, 4, ...), a logarithmic scale spaces values exponentially (1, 10, 100, 1000, ...).
-   The distance between ticks represents a multiplicative factor instead of an additive one.

```{r}
#| echo: FALSE
#| fig-width: 12
#| fig-height: 6

### ALL OF THIS IS SUPER SPICY 🌶️🌶️️ and beyond what I would expect of you in this course...

library(scales)

# Function to generate breaks for log10 scale
minor_log_breaks <- function() {
  as.numeric(outer(1:10, 10^(0:6)))  # Generates minor breaks at 1.1, 1.2, ..., 9, 10, 20, 30...
}

# Define major log breaks for both plots (powers of 10)
major_log_breaks <- 10^(0:6)  # 1, 10, 100, 1000, ..., 1,000,000

# Define linear breaks for both plots (powers of 10)
major_linear_breaks <- seq(0, 30000000, 5000000)
minor_linear_breaks <- seq(0, 30000000, 1000000)

# Scatterplot with Linear Scale
p1 <- ggplot(can_lang, aes(x = most_at_home, y = mother_tongue)) +
  geom_point() +
  scale_x_continuous(breaks = major_linear_breaks, 
                     labels = label_comma(), 
                     minor_breaks = minor_linear_breaks) +
  scale_y_continuous(breaks = major_linear_breaks,
                     labels = label_comma(), 
                     minor_breaks = minor_linear_breaks) +
  labs(
    x = "Language spoken most at home \n (number of Canadian residents)",
    y = "Mother tongue \n (number of Canadian residents)",
    title = "Linear Scale on x- and y-axis", 
    subtitle = "the next break on a linear scale can be found my ADDING"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_line(color = "lightblue", linetype = "solid"), # Light blue for major grid lines
    panel.grid.minor = element_line(color = "#D6EBF2", linetype = "solid")
  )

# Extract breaks from the linear scale plot
p1_build <- ggplot_build(p1)
extracted_linear_breaks <- p1_build$layout$panel_params[[1]]$x$breaks

# Remove zeros (log10(0) is undefined)
extracted_linear_breaks <- extracted_linear_breaks[extracted_linear_breaks > 0]

# Transform breaks for log scale
log_breaks <- 10^floor(log10(extracted_linear_breaks))  # Round to nearest power of 10

# Custom function to format numbers as scientific notation (10^x)
format_scientific <- function(x) {
  parse(text = paste0("10^", log10(x)))  # Converts numbers into properly formatted exponents
}

# Scatterplot with Log Scale
p2 <- ggplot(can_lang, aes(x = most_at_home, y = mother_tongue)) +
  geom_point() +
  scale_x_log10(
    minor_breaks = minor_log_breaks(),
    breaks = major_log_breaks,  # Use manually defined major breaks
    labels = label_comma(),
    sec.axis = sec_axis(~ .,  
                        breaks = major_log_breaks,  
                        labels = format_scientific,  # Scientific notation labels
                        name = NULL)  
  ) +
  scale_y_log10(
    minor_breaks = minor_log_breaks(),
    breaks = major_log_breaks,  
    labels = label_comma(),
    sec.axis = sec_axis(~ .,  
                        breaks = major_log_breaks,  
                        labels = format_scientific,  # Scientific notation labels
                        name = NULL)  
  ) +
  labs(
    x = "Language spoken most at home \n (number of Canadian residents, log scale)",
    y = "Mother tongue \n (number of Canadian residents, log scale)",
    title = "Log Scale on x- and y-axis (log-log scale)", 
    subtitle = "The next break on a log scale can be found by MULTIPLYING"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_line(color = "lightblue", linetype = "solid"),  # Light blue for minor grid lines
    panel.grid.major = element_line(color = "darkred", linetype = "dotted"),  # Dark red dotted for major grid lines
    axis.title.x.bottom = element_text(margin = margin(t = 10)),
    axis.text.x.top = element_text(margin = margin(b = 5)),
    axis.title.y.left = element_text(margin = margin(r = 10)),  # Add space for primary y-axis
    axis.text.y.right = element_text(margin = margin(l = 5))  # Add space for secondary y-axis
  )


# Display both plots side by side
library(gridExtra)
grid.arrange(p1, p2, ncol = 2)

```

See how much more clearly we can see all the points!
:::

For you to do this yourself, you need to use `scale_*_log10()` instead of `scale_*_continuous()`:

```{r}
# same plot, change x and y to scale_*_log10
# set labels=label_comma() so that we still get our commas
can_lang_baseplot +
  scale_x_log10(labels=label_comma()) +
  scale_y_log10(labels=label_comma())


```

1.  `scale_x_log10()` converts x-axis to a log-scale
2.  `scale_y_log10()` converts y-axis to a log-scale


::: callout-tip
# Use ✅ `scale_*_log10()` instead of 🚫`log(variable)`

```{r}
#| echo: FALSE

# Scatterplot with Log Scale (Recommended)
p1 <- ggplot(can_lang, aes(x = most_at_home, y = mother_tongue)) +
  geom_point() +
  scale_x_log10(labels = label_comma()) + 
  scale_y_log10(labels = label_comma()) + 
  labs(
    title = "Recommended Approach", 
    subtitle = "Use `scale_*_log10()`",
    x =  "Language spoken most at home \n (number of Canadian residents)",
    y = "Mother tongue \n (number of Canadian residents)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(color = "darkgreen", size = 12, face = "bold"), 
    text = element_text(family = "sans")  # Ensures emoji support in most environments
  )

# Scatterplot with Manual Log Transform (Not Recommended)
p2 <- ggplot(can_lang, aes(x = log(most_at_home), y = log(mother_tongue))) +
  geom_point() +
  scale_x_continuous(labels = label_comma()) + 
  scale_y_continuous(labels = label_comma()) + 
  labs(
    title = "Not recommended", 
    subtitle="Manually Transforming the Data (log(variable))",
    x =  "Language spoken most at home \n (number of Canadian residents)",
    y = "Mother tongue \n (number of Canadian residents)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(color = "red", size = 12, face = "bold"), 
    text = element_text(family = "sans")  
  )

# Display both plots side by side
gridExtra::grid.arrange(p1, p2, ncol = 2)
```
:::


## Labels

### Using `geom_text`

Adding text to a plot is one of the most common forms of annotation. Most plots will not benefit from adding text to every single observation on the plot, but labeling outliers and other important points is very useful.

A add label for each language in this dataset using `geom_text(aes(label = language))`:

```{r}

# same base plot, use scale_x_log10 
# set labels=label_comma()

can_lang_baseplot +
  geom_text(aes(label=language)) +
  scale_x_log10(labels=label_comma()) +
  scale_y_log10(labels=label_comma())



```

Yikes! This is way too much going on in one plot. A few options to try when this happens:

-   Decrease the font size of the labels (using the `size=` argument inside `geom_text`).
-   Use the `ggrepel` package to spread out the labels a bit more
-   Pick out only a subset of the points to label


### Using `ggrepel`

```{r}
# install and load the ggrepel library
library(ggrepel)

# add labels using geom_text_repel assign label=language in aes()

can_lang_baseplot +
  geom_text_repel(aes(label=language)) +
  scale_x_log10(labels=label_comma()) +
  scale_y_log10(labels=label_comma())

# use max.overlaps argument to display more labels
can_lang_baseplot +
  geom_text_repel(aes(label=language),
                  max.overlaps=25) +
  scale_x_log10(labels=label_comma()) +
  scale_y_log10(labels=label_comma())

```

### Subset the labels

Create a new column for the labels. Use `case_when` to only use the official language names and not to put a label for other language categories.

```{r}
# mutate new column for official languages using case_when()
# This column will have the name of the language only when it is an Official Language and otherwise contain NAs

# remember the second half if you are are assigning a value to the things you have not set an explicit rule for you use TRUE ~ NA

can_lang |>
  distinct(category)

can_lang <- can_lang |>
  mutate(official_lang = case_when(
    category=="Official languages" ~ language, 
    TRUE ~ NA)) 

```

```{r}
# now make plot, using new base plot with just the official languages labeled

# run code with old base plot
# commented out because it is broken
# need to rerun the baseplot(below) with the updated can_lang
# can_lang_baseplot +
#   geom_text_repel(aes(label=official_lang)) +
#   scale_x_log10(labels=label_comma()) +
#   scale_y_log10(labels=label_comma())

# We need to redo the base plot with the new can_lang dataset with the new official_languages column in it 

can_lang_baseplot <- can_lang |>
  ggplot(aes(x=most_at_home, y= mother_tongue)) +
  geom_point() +
  xlab("Language spoken most at home \n (number of Canadian residents)") +
  ylab("Mother tongue \n (number of Canadian residents)") +
  theme_tufte()

# Try the plot again

can_lang_baseplot +
  geom_text_repel(aes(label=official_lang)) +
  scale_x_log10(labels=label_comma()) +
  scale_y_log10(labels=label_comma())

# play around with color

can_lang_baseplot +
  geom_text_repel(aes(label=official_lang), color="red") +
  scale_x_log10(labels=label_comma()) +
  scale_y_log10(labels=label_comma())

```

## Facet Wrap

`facet_wrap()` is a function in the `ggplot2` package that allows you to create a multi-panel plot showing a similar plot over different subsets of the data, usually different values of a categorical variable.

Create separate side-by-side plots for each different category of language.

```{r}
# make the baseplot with facets by category
can_lang_baseplot +
  geom_text_repel(aes(label=official_lang), color="red") +
  scale_x_log10(labels=label_comma()) +
  scale_y_log10(labels=label_comma()) +
  facet_wrap(~category)


# make the same plot with different themes
can_lang_baseplot +
  geom_text_repel(aes(label=official_lang), color="red") +
  scale_x_log10(labels=label_comma()) +
  scale_y_log10(labels=label_comma()) +
  facet_wrap(~category) +
  theme_minimal()

can_lang_baseplot +
  geom_text_repel(aes(label=official_lang), color="red") +
  scale_x_log10(labels=label_comma()) +
  scale_y_log10(labels=label_comma()) +
  facet_wrap(~category) +
  theme_bw()

can_lang_baseplot +
  geom_text_repel(aes(label=official_lang), color="red") +
  scale_x_log10(labels=label_comma()) +
  scale_y_log10(labels=label_comma()) +
  facet_wrap(~category) +
  theme_economist()


# set scales="free" to show how the axes can be altered
can_lang_baseplot +
  geom_text_repel(aes(label=official_lang), color="red") +
  scale_x_log10(labels=label_comma()) +
  scale_y_log10(labels=label_comma()) +
  facet_wrap(~category, scales="free")


```

### Make a more complicated faceted plot

In class we talked about how to add colored points on top of the full data set with the facets. Here is a cleaned up version of the solution that I came up with. I also poked around online and could not come up with a fundamentally different way of doing this. 

```{r}
# create a data set that does not contain the category column
can_lang_abbr <- can_lang |>
  select(mother_tongue, most_at_home)

ggplot() +
  geom_point(data=can_lang_abbr, #use abbreviated data set for the background points
             aes(x=most_at_home, y=mother_tongue), color="gray") +
  geom_point(data=can_lang, # use the full data set 
             aes(x=most_at_home, 
                 y=mother_tongue, 
                 color=category)) + # map category to color
  scale_color_manual(values=c("#E1AF00","#3B9AB2", "#F21A00")) +
  scale_x_log10(labels=label_comma()) +
  scale_y_log10(labels=label_comma()) +
  xlab("Language spoken most at home \n (number of Canadian residents)") +
  ylab("Mother tongue \n (number of Canadian residents)") +
  facet_wrap(~category) + #facet by category
  theme_tufte() +
  theme(legend.position = "none")
  
```

