---
title: "STAT 118: Homework G"
subtitle: 'Maps with `maps` and `sf`'
author: "YOUR NAME HERE"
format:
  html: 
    toc: true
    code-overflow: wrap
    code-fold: true
    embed-resources: true
---

```{r message=FALSE, warning=FALSE}
#LOAD PACKAGES 
library(tidyverse)
library(sf) #this is a package needed so R can work with sf objects
```

```{r message=FALSE, warning=FALSE}
#LOAD DATA
library(spData) #this packages contains the dataset (with sf objects) that we will be using today
data("us_states")

library(rnaturalearth)
library(rnaturalearthdata)
world <- ne_countries(scale = "medium", returnclass = "sf")
africa <- world[world$continent == "Africa", ]

```

# 1.
Create a basic map of the countries of the world. To access `sf` geometries for the world use the `world` dataset loaded above. The land should be some shade of green. Use something other than the default ggplot theme. 




# 2. 

Use the columns `pop_est` and `gdp_md` to create a new column called `gdp_per_cap` which contains the GDP per capita for each country in the world in this dataset. Print out the first few rows of your dataset to confirm that you have succeeded. 



# 3.
Create a choropleth map which colors each country depending on the GDP per capita. 

- Utilize an appropriate color palette for this data
- Add a theme
- make sure the legend labels are clear to a general audience (you might need to research this a bit...)
- the country borders should be "black"

*Your figure should not contain any negative values. And, Antarctica should not have the highest GDP per capita in the world. You should diagnose these issues and replace the clearly incorrect values with NAs.*



# 4.
Create a choropleth map which colors each country based on their economy type (`economy`). 

- Utilize an appropriate color palette for this data
- Add a theme
- make sure the legend labels are clear to a general audience (you might need to research this a bit...)
- the country borders should be "grey"



# 5. 

Create a choropleth map of Africa based on the population in each country. 
Use `scale_fill_viridis_c(option = "viridis")`. 




# 6. 

Create a choropleth map of Africa based on the population in each country. 
Use `scale_fill_viridis_c(option = "viridis", trans = "log")`. 



# 7. 

What does the `trans="log"` argument do? Why was it applied? 


# 8. 

Add country names to your map from #6. 





# Exploring the Environmental Justice Index

Open up you Midterm Assignment on Canvas. Use the links in the Midterm Assignment to access the information referenced below.

### 1. Get to know the data. 

With a partner, use the EJI website and data dictionary to figure out:
   - What does one row represent?
   - What kinds of information are included?
   - How should you interpret a high EJI value?
   - Find 2–3 variables that interest you.
   
Write a 3–4 sentence description of the dataset in your own words.
   
### 2. Open and subset the data. 

Create a new object called `eji_vt` that just contains data for Vermont.

3. Make a map. 

Choose one variable that interests you and make a choropleth map of it for Vermont. Use an appropriate color scale and clearly label your legend.

4. Explore a relationship

Choose a second EJI variable that you think might be related to the variable you mapped. Create a scatterplot comparing the two variables. What relationship do you see?

5. Ask a question

Based on your exploration, write down one question you would be interested in investigating further with the EJI dataset.
