---
title: '`forcats`: working with categorical data'
author: 'Casey Schine adapted from Emily Malcolm-White'
format:
  html: 
    toc: TRUE
    code-overflow: wrap
    embed-resources: true
    code-tools:
      source: true
      toggle: false
      caption: none
execute: 
  message: FALSE
  warning: FALSE
---

The R package `forcats` is designed to make working with categorical variables easier and more efficient. It provides a set of functions that allow you to manipulate and analyze categorical data with ease. In this lesson, we'll cover the basics of the `forcats` package and some of its most useful functions.

## Categorical Variables

Let's review what categorical data is. Categorical data is a type of data that consists of categories or labels.

Examples of categorical data include:

-   Colors (red, blue, green, etc.)
-   Types of vehicles (sedan, SUV, truck)
-   Educational degrees (high school, college, graduate school)

Categorical data can be further divided into two types: *nominal* and *ordinal*. Nominal data consists of categories that have no inherent order, while ordinal data consists of categories that have a natural order. For example, educational degrees are ordinal data because they can be ordered from least to most advanced.

## `mpg` Data

We will play with different functions in the `forcats` packages using the `mpg` dataset from earlier in the semester.

```{r}
library(forcats)
library(tidyverse)
data("mpg")
library(ggthemes)
```

Let's take a look at a side-by-side boxplot:

```{r}
mpg |> 
  ggplot(aes(x=class, y=hwy)) +
  geom_boxplot() +
  theme_tufte()
```

## Reordering Factor Levels

One of the most useful functions is fct_relevel(), which allows you to reorder the levels of a factor. This can be useful when you want to change the default ordering of the levels or when you want to group certain levels together.

Is `class` a factor?

```{r}
# use is.factor() to inquire if class is a factor variable
# returns TRUE or FALSE

# with dplyr syntax (pipes)
# with select()
mpg |> select(class) |> is.factor()

# with pull()
mpg |> pull(class) |> is.factor()

```

Let's make it a factor!

```{r}
# use mutate to write over class with new factor variable called class
# use as.factor()

mpg <- mpg |>
  mutate(class = as.factor(class))
```

Let's check the levels and their current ordering!

```{r}

# use levels() to get factor levels
# levels() requires vector so must use pull() rather than select()


mpg |> 
  pull(class) |> 
  levels()


```

To reorder the levels with `fct_relevel()`

```{r}
mpg <- mpg |> 
  mutate(class = class |> fct_relevel( "compact","subcompact","midsize","2seater","minivan","suv","pickup"))

mpg |> 
  pull(class) |> 
  levels()
```

Let's recreate our side-by-side boxplot now:

```{r}
mpg |>
  ggplot(aes(x=class, y=hwy)) +
  geom_boxplot(color="magenta", fill="linen") +
  theme_tufte()
```

Rather than reordering them manually by typing the order, you could also re-level by some numeric criteria using `fct_reorder()`. For example:

```{r}
#?fct_reorder

# use pipe within mutate as class |> fct_reorder(hwy, median)
mpg <- mpg |>
  mutate(class = class |> fct_reorder(hwy, median))

# do the same thing without the pipe as fct_reorder(class, hwy, median)
mpg <- mpg |>
  mutate(class = fct_reorder(class, hwy, median))

# take a look at the order again
mpg |> 
  pull(class) |> 
  levels()

# remake the boxplot again with the new order
mpg |>
  ggplot(aes(x=class, y=hwy)) +
  geom_boxplot(color="magenta", fill="linen") +
  theme_tufte()

# what other functions can we use to reorder???

```

## Renaming Factor levels with `fct_recode`

Sometimes you might not like the way the levels are named.

```{r}
# use fct_recode to change the name of a factor level in class
# fct_recode("two-seater" = "2seater")
## NEW NAME = OLD NAME

mpg <- mpg |> 
  mutate(class = class |>  fct_recode("two-seater" = "2seater"))

# take another look at the factor
mpg |> 
  pull(class) |>
  levels()

```

## Factor Collapsing with `fct_collapse()`

Let's say we wanted to create only two categories -- cars and larger vehicles.

```{r}

# use fct_collapse to collapse class into cars and big factors
# mutate on new variable called class_two
# fct_collapse( cars = c("compact", "subcompact", "midsize", "two-seater"), big = c("pickup", "suv", "minivan"))
mpg <- mpg |>
  mutate(class_two = class |>  fct_collapse( cars = c("compact", "subcompact", "midsize", "two-seater"), big = c("pickup", "suv", "minivan")))

# take a look at the new variable factor levels
mpg |> 
  pull(class_two) |>
  levels()
```

## Lumping into an other category

-   `fct_lump_min()`: lumps levels that appear fewer than min times.

-   `fct_lump_prop()`: lumps levels that appear in fewer than (or equal to) prop \* n times.

-   `fct_lump_n()` lumps all levels except for the n most frequent (or least frequent if n \< 0)

```{r}
# count manufacturers in dataset
mpg |> 
  count(manufacturer)
```



Let's say we wanted only the manufacturers with at least 15 cars produced. Everything else we want to just be other:

```{r}

# make manufacturer lump variable using fct_lump_min(15)
mpg <- mpg |>
  mutate(man_lumped = manufacturer |> fct_lump_min(15))

# look at the result with count()


# now with fct_lump_n() for the top 5

# look at the result with count()


# now with fct_lump_prop() for 0.1 and -0.05
mpg |>
  mutate(man_lumped = manufacturer |> fct_lump_prop(-0.05)) |>
  count(man_lumped)

# look at the result with count()



```

## Kable Tables

Create a table using kableExtra:

```{r}
library(kableExtra)

# make table for manufacturer before and after lumping 
# using kbl() |> kable_styling()

mpg |>
  count(manufacturer) |>
  kbl() |>
  kable_styling()

mpg |>
  count(man_lumped) |>
  kbl() |>
  kable_styling()


```
By default, `count()` puts the variable in descending order (alphabetically). If we want to change the order we can use some of the tricks that we just learned for reordering factors! 

```{r}
## Tables + factors: controlling order and labels

# make a new table using count(class) variable
# then |> kbl() |> kable_styling()

mpg |>
  count(class) |>
  kbl() |>
  kable_styling()

# you can reorder the factor before feeding it into the table to control the order
# let's go back to our first reordering
# fct_relevel( "compact","subcompact","midsize","2seater","minivan","suv","pickup"))

# then add a caption in the kbl() function as caption="caption goes here"
mpg |>
  mutate(class = fct_relevel(class,
                             "compact","subcompact","midsize","two-seater",
                             "minivan","suv","pickup")) |>
  count(class) |>
  kbl(caption = "Counts by vehicle class (ordered by factor levels)") |>
  kable_styling()


# what if we want to show something other than count()
# we can use group_by() and summarize() to get statistics for each group
mpg |>
  group_by(class) |>
  summarize(median_hwy=median(hwy)) |>
  kbl() |>
  kable_styling()

# we can also use the fct_reorder() function to reorder the factor by a calculated value (as we did above)
# class = fct_reorder(class, hwy, .fun = median)

mpg |>
  mutate(class = fct_reorder(class, hwy, .fun = median)) |>
  group_by(class) |>
  summarize(
    n = n(),
    median_hwy = median(hwy),
    .groups = "drop"
  ) |>
  arrange(class) |>
  kbl(digits = 1,
      caption = "Vehicle class ordered by median highway MPG") |>
  kable_styling()
```

We can also control aspects of the table and how it looks using arguments in the `kbl()` and `kable_styling()` functions. 

```{r}

# how can we control aspects of the table 
#?kbl to look at some of the options

# can set the number of digits using the digits=x argument in kbl()

# also col.names argument to change the column names for the output table


mpg |>
  group_by(class) |>
  summarize(median_hwy=median(hwy)) |>
  kbl(digits=0, col.names=c("Class", "Median Hwy MPG")) |>
  kable_styling()

```

