---
title: 'working with text with `stringr`'
author: 'Casey Schine adapted from Violet Ross'
format:
  html: 
    toc: TRUE
    code-overflow: wrap
    embed-resources: true
editor_options: 
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE}
#LOAD PACKAGES 
library(tidyverse)
library(stringr)
```

## A few basics

**What is a string?**

A string is a data type used to represent text. In R, we indicate that something is a string by referring to it with quotes.

Examples of strings:

-   Rwanda
-   #11111
-   1w3r
-   1999_1

NOT a string:

-   1999

**What is** `stringr` **?**

stringr is a package containing a bunch of functions that help us work with strings.

[**stringr cheat sheet**](https://evoldyn.gitlab.io/evomics-2018/ref-sheets/R_strings.pdf)

# Using `stringr`

**I'm registering for classes for this Fall and am trying to decide which courses to take.**

Let's look at the course catalog!

Read in the courses data.

```{r, message=FALSE, warning=FALSE}
# Download the courses.csv dataset

# you will need to put in your own filepath here
# !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
courses <- read_csv("~/Dropbox/!Middlebury/Teaching/STAT 118/S2026/S2026_R_Proj/data/courses.csv")
# !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!

view(courses)

```

## `str_detect`

**inputs**:

-   a string (or a column of strings) to search *from*\
-   a string to search *for*

**output**: boolean value (TRUE or FALSE)

little example:

```{r}
# using str_detect to search for strings
# for the string "Welcome to data science, look at this cool data"

#find the pattern "data" 
str_detect("Welcome to data science, look at this cool data", "data")

# find the pattern "pineapple" 
str_detect("Welcome to data science, look at this cool data", "pineapple")
```

```{r}

# note for me to look into
# can you do this for a whole dataset??? analog for the lookfor function in Stata

# to select columns that contain a specific string:
courses |>
select(contains("course"))

# you can also use ends_with(), starts_with(), and matches()


# this code will filter for all rows that contain the pattern "CHEM" no matter where it appears
courses |>
  filter(if_any(everything(), ~str_detect(.x, "CHEM")))

```

I only want to take classes in Warner! Let's filter for rows where we **detect** the word "Warner" in the `location` column.

```{r}
# filter for where the location column contains the pattern "WNS"

courses |>
  distinct(location) |>
  pull()

warner_classes <- courses |>
  filter(str_detect(location, "WNS"))

warner_classes

# look at what str_detect actually outputs in column
warner_classes <- courses |>
  mutate(loc_boolean=str_detect(location, "WNS"))

warner_classes

```

Suppose I don't want any classes on Friday. Let's use `str_detect` to find our options.

```{r}
# look for instances of the pattern "Friday" in the meet column
# using str_detect() 
# filter them out

nofriday_classes <- courses |>
  filter(!str_detect(meet, "Friday"))

nofriday_classes

# we can do the same thing using the negate argument in str_detect
nofriday_classes <- courses |>
  filter(str_detect(meet, "Friday", negate=TRUE))

nofriday_classes

# this doesn't work because the string is not an exact match due to difference in case
nofriday_classes <- courses |>
  filter(!str_detect(meet, "friday"))
```

Perhaps I'm interested in immigration. We can use `str_detect` to find course descriptions that mention immigration.

The `regex` function is used to write regular expressions in R. Regular expressions are helpful if you want to search for a pattern rather than a specific word or phrase.\
For now, we will only use regex to ignore capitalization. If you're interested in using regular expressions at some point, this [regex cheat sheet](https://cheatography.com/davechild/cheat-sheets/regular-expressions/) will be super helpful.

```{r}

# filter for class descriptions that contain the pattern "immigration" using str_detect()
# use regex to ignore upper versus lower case as: regex("immigration", ignore_case=TRUE)

immigration_courses <- courses |>
  filter(str_detect(description, regex("immigration", ignore_case=TRUE)))

immigration_courses

```

## `str_extract` and `str_remove`

**str_extract inputs**:

-   a string (or a column of strings) to extract *from*\
-   a string to extract

**str_extract output**: the string you wanted to extract OR nothing

**str_remove inputs**:

-   a string (or a column of strings) to remove stuff *from*\
-   a string to remove

**str_remove output**: the first input string with the second input removed

little example:

```{r}
# for the string "Welcome to data science, look at this cool data"
# and the pattern "data"

# str_extract() extracts first instance
str_extract("Welcome to data science, look at this cool data", "data")

# str_extract_all() extracts all instances
str_extract_all("Welcome to data science, look at this cool data", "data")

# str_remove() removes first instance
str_remove("Welcome to data science, look at this cool data", "data")

# str_remove_all() removes all instances
str_remove_all("Welcome to data science, look at this cool data", "data")

```

CW is part of the distribution requirement column. I want CW to be its own column.\
First we can create a new column called `isCW` by **extracting** instances of "CW" from the `distros` column in each row.\
Then we can **remove** all instances of "CW" from the `distros` column.

```{r}

# make new column that states whether a course is CW or not
# and count how many classes are CW

courses_CW <- courses |>
  mutate(isCW = str_extract(distros, "CW"))

courses_CW

# now make a new column for distros and remove CW using str_remove()
courses_CW <- courses |>
  mutate(isCW = str_extract(distros, "CW")) |>
  mutate(distros = str_remove(distros, "CW"))

courses_CW
  
# count the number of CW courses
courses_CW |>
  count(isCW)

```

## `str_sub`

**str_sub inputs**:

-   a string (or a column of strings) take substrings of\
-   where the substring should start
-   where the substring should end

**str_sub output**: the substring

little example:

```{r}

#extract the pieces of a string using str_sub() function
# extract the substring data science
# from the sentence "Welcome to data science, look at this cool data"
# need to define the position of the start character and the end character

str_sub("Welcome to data science, look at this cool data",
        start=12, end=23)


# move the start and end to extract a different string
# the default end is the end of the string
str_sub("Welcome to data science, look at this cool data",
        start=39)
```

Maybe I only want 200 level math and stats classes.\
First we filter for just math/stats classes. Then we can create a new column called `level` that contains only the sixth character from the `courses` column.

We can then filter our `level` column for specific course levels.

```{r}
# first let's get all of the exact department names
courses |>
  distinct(department) |>
  pull()

# mutate on a column called level that extracts the 6th character from courseNum column

Math_200 <- courses |>
  filter(str_detect(department, "Mathematics")) |>
  mutate(level=str_sub(courseNum , start=6, end=6)) |>
  filter(level==2) #filter for level 2 only

Math_200

# filter for 200 level classes without mutating first

Math_200 <- courses |>
  filter(str_detect(department, "Mathematics")) |>
  filter(str_sub(courseNum, start=6, end=6)==2)

Math_200

```

## `str_count`

**str_count inputs**:

-   a string (or a column of strings) to count *from*\
-   a string you want to count the number of instances of

**str_count output**: the number of times the second string occurs in your first string

little example:

```{r}
# count the instance of the word "data" in the sentence "Welcome to data science, look at this cool data"
# using str_count()

str_count("Welcome to data science, look at this cool data", "data")

str_count("Welcome to data science, look at this cool data", "look")
```

Maybe I only want my classes to meet twice a week. Let's **count** how many times "day" appears in the `meet` column and put this value in a new column called `day_count`. `day_count` represents how many days a week a class meets.

```{r}

# mutate on a column called day_count using str_count() for the meet column, looking for instances of the string "day"

# save day_count as a column 

course_days <- courses |>
  mutate(day_count=(str_count(meet, "day")))


# now extract what is the max number of days a class meets

course_days |>
  summarize(max_days=max(day_count)) |>
  pull()

# which courses meet 6 days a week
course_days |>
  filter(day_count==6) 

```

Let's visualize this data.

```{r}
# make a bar plot of dayCount with x=day_count 

course_days |>
  ggplot() +
  geom_bar(aes(x=day_count)) +
  theme_minimal()


  
```

## `word`

Allows you to select words from a string. The function assumes as a default that the words are separated by a space, but you can define a different separator if you need using the `sep=` argument.

```{r}
# let's look at the help docs
#?word


word("Welcome to data science, look at this cool data", 3)

# mutate new column called building with just the first word from location

course_building <- courses |>
  mutate(building = word(location, start=1, end=2))
```

Let's make a plot using some of the columns we've just made:

```{r}
# what should we plot???

# count the number of words in the description
# we do this by counting the number of spaces and then adding 1
course_description <- courses |>
  mutate(desc_words = str_count(description, pattern=" ") + 1)

course_description |>
  ggplot(aes(x = desc_words)) +
  geom_histogram(binwidth = 10) +
  theme_minimal() +
  labs(
    x = "Number of words in description",
    y = "Number of courses",
    title = "How long are course descriptions?"
  )

# Now I'm wondering which courses have the longest descriptions
course_description |>
  filter(desc_words>150) |>
  view()

# Now I want to know how the word count distribution changes by department
library(ggridges)

course_description |>
  mutate(department=factor(department),
         department=fct_lump_min(department,10),
         department=fct_reorder(department, desc_words, .fun=max)) |>
  filter(department != "Other") |>
  ggplot(aes(x = desc_words, y = department)) +
  geom_density_ridges(aes(fill=department)) +
  theme_minimal() +
  labs(x = "Number of words in description",
       y = "Department",
       title = "How long are course descriptions by department?") +
  theme(legend.position = "none")

# I had to stop myself here

```

## In groups

Make a bar plot with the number of classes that meet in each building. Put the building name on the y-axis and the number of classes that meet there on the x-axis. Put the y-axis in order from the building with the most classes to the building with the least classes. It doesn't have to be perfect. With more time and for a more polished plot there are things you would do to clean it up. This is about combining some of the tools you just learned with some that you learned throughout the semester.

```{r}
# mutate column called building with first 2 words from location 

course_building <- courses |>
  mutate(building = factor(word(location, start=1, end=2)))

course_building

course_building |>
  filter(location=="bleh") 
```

```{r}
# make bar plot of the number of classes by building 

course_building |>
  group_by(building) |>
  summarize(count=n()) |>
  mutate(building=fct_reorder(building, count)) |>
  ggplot() +
  geom_col(aes(y=building, x=count)) +
  theme_minimal()


```
