---
title:  'Making plots  with `ggplot2`: Barplots and Scatterplots'
author: 'Emily Malcolm-White'
format:
  html: 
    toc: TRUE
    code-overflow: wrap
    embed-resources: true
execute: 
  message: FALSE
  warning: FALSE
---

`ggplot2` is a package built within the `tidyverse` package for creating awesome graphs!

```{r}
library(tidyverse)
```

```{r}
#Import the can_lang dataset 
can_lang <- read_csv("https://raw.githubusercontent.com/ttimbers/canlang/master/inst/extdata/can_lang.csv")
```

# Recall our Top 10 example:

This code gave a list of 10 Aboriginal Languages which have the most number of people who speak them as their mother tongue:

```{r}
ten_lang <- can_lang |> 
  filter(category == "Aboriginal languages") |>
  arrange(desc(mother_tongue)) |>
  select(language, mother_tongue) |>  
  slice(1:10) #
```

# Barplots

Suppose we wanted to display this information in a barplot instead of in a table.

```{r}
# start with basic ggplot
# data and aes()
# plus instead of pipe
# geom_col()

ggplot(data=ten_lang, aes(x = language, y = mother_tongue)) + #plus instead of pipe
  geom_col()
 
```

Is there any improvements we could make to this graph?

# To better view text

Display the bars horizontally instead of vertically!

```{r}

ggplot(data=ten_lang, aes(y = language, x = mother_tongue)) + #plus instead of pipe
  geom_col()

```

1.  This flips the x and y axes!

# Colors, Shapes, and Sizes

```{r}
ggplot(ten_lang, aes(x = mother_tongue, y = language)) +
  geom_bar(stat = "identity") 
```

::: callout-tip
Barplots are good for displaying one categorical variable and one numeric variable. The number variable could be counts (as above) or they could be averages or totals or maximums or minimums (or many other things!)
:::

# ggplot: scatterplot with `geom_point`

::: callout-tip
Scatterplots are good for displaying the relationship between two numerical variables.
:::

The `mtcars` dataset was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models). It's available inside the `ggplot` package which is already installed.

```{r}
#load the data
data(mtcars)
```

```{r}

```

Note that you can change the color, shape (`pch` for plotting character) and size of these points!

### Quick update to the dataset

```{r}
#code to update `mtcars` dataset so that `am` is treated as a factor rather than a continuous numeric variable
mtcars <- mtcars %>%  
  mutate(am = as.factor(am)) 
```

This modifies the am column, which represents the transmission type of the car (0 = automatic, 1 = manual). The as.factor(am) function converts the am variable from a numeric type (0 or 1) into a categorical factor.

# Inside `aes()` or outside `aes()`?

What is the difference between these two graphs?

```{r}
# set color outside aesthetics
ggplot(mtcars, aes(x=wt, y=mpg)) +
  geom_point(color="red") 
```

```{r}
# set color in aesthetics
ggplot(mtcars, aes(x=wt, y=mpg)) +
  geom_point(aes(color=am)) #<1>
```

Now let's change the shape of the points to designate the different groups.

```{r}
#color in aesthetics and then customize with scale_color_manual()
ggplot(mtcars, aes(x=wt, y=mpg)) +
  geom_point(aes(shape=am)) 
```

We can add back in color:

```{r}
#color in aesthetics and then customize with scale_color_manual()
ggplot(mtcars, aes(x=wt, y=mpg)) +
  geom_point(aes(shape=am, color=am)) 
```

And we can add a size argument as well:

```{r}
#color in aesthetics and then customize with scale_color_manual()
ggplot(mtcars, aes(x=wt, y=mpg)) +
  geom_point(aes(shape=am, color=am, size=hp)) 
```

::: callout-tip
-   If the thing you are trying to change (color, shape, size, etc.) depends on a variable, you should put in *inside* the aesthetics
-   If the thing you are trying to change (color, shape, size, etc.) should happen for all things, you should not put it inside the aesthetics.
:::

# Customizing Colors in Aesthetics

```{r}
#color in aesthetics and then customize with scale_color_manual()
ggplot(mtcars, aes(x=wt, y=mpg)) +
  geom_point(aes(color=am)) +
  scale_color_manual(values=c("black", "orange")) 
```

# Global vs. Local Aesthetics

Global aesthetic mappings apply to all geometries and can be defined when you initially call `ggplot()`. All the geometries added as layers will default to this mapping. Local aesthetic mappings add additional information or override the default mappings.

```{r}
#set color = am as a global aesthetic in ggplot()
ggplot(mtcars, aes(x=wt, y=mpg, color=am)) +
  geom_point()
```

```{r}
#set color = am as a local aesthetic in geom_point
ggplot(mtcars, aes(x=wt, y=mpg)) +
  geom_point(aes(color=am))
```

```{r}
# set color = am as global aesthetic in ggplot()
# then overwrite color = am with a local aesthetic in geom_point
# do it first with empty geom_point, then define color
ggplot(mtcars, aes(x=wt, y=mpg, color=am)) +
  geom_point(color="purple")
```

# Piping in Data

You can define your dataset within ggplot as we have been doing or you can pipe in a wrangled version of your data.

Here is our basic figure, with our dataset defined within `ggplot()`:

```{r}

ggplot(mtcars, aes(x=wt, y=mpg)) +
  geom_point(aes(color=am))
```

If we pipe in the same data set, we get the same plot:

```{r}
mtcars |>
  ggplot(aes(x=wt, y=mpg)) +
  geom_point(aes(color=am))

```

But piping in the data allows us to use our wrangling skills to send in a slightly (or sometimes very) different dataset without having to create a new object. Here is the same plot, but with am filtered for only the 0 values:

```{r}
mtcars |>
  filter(am==0) |>
  ggplot(aes(x=wt, y=mpg)) +
  geom_point(aes(color=am))
```

# Quick Introduction to Themes

There are these things in `ggplot()` called themes that change the look of a figure. Here are a few examples:

```{r}
#color in aesthetics and then customize with scale_color_manual()
ggplot(mtcars, aes(x=wt, y=mpg)) +
  geom_point(aes(shape=am, color=am, size=hp)) +
  theme_bw()

ggplot(mtcars, aes(x=wt, y=mpg)) +
  geom_point(aes(shape=am, color=am, size=hp)) +
  theme_minimal()

ggplot(mtcars, aes(x=wt, y=mpg)) +
  geom_point(aes(shape=am, color=am, size=hp)) +
  theme_dark()
```
