---
title: 'Working with dates using `lubridate`'
author: 'Casey Schine adapted from Emily Malcolm-White'
format:
  html: 
    toc: TRUE
    code-overflow: wrap
    embed-resources: true
    code-tools:
      source: true
      toggle: false
      caption: none
    code-annotations: hover
  pdf: default
execute: 
  message: FALSE
  warning: FALSE
editor_options: 
  chunk_output_type: console
---

Dates and times are everywhere in data: timestamps on social media posts, transaction dates in sales records, birthdates in survey data. But unlike numbers or strings, dates are tricky—you can’t just subtract them like normal numbers, and parsing them from messy formats can be a headache.

# Date Formats

Think of how many different formats you know of to format a date:

-   

-   

Yikes!

# Date, Time, and Datetime

Date/time data are data that conveys information about, you guessed it, date and/or time! There are three relevant data types when we talk about date/time data:

1.  Date - only has the date (e.g. 2020-05-15)
2.  Time - only has the time (e.g. 20:45:00)
3.  Datetime - has both the date and time (e.g. 2020-05-15 20:45:00)

# `Lubridate`

```{r}
#LOAD PACKAGES 
library(tidyverse) #lubridate is in the tidyverse!
```

# Standard Date Format

The `ymd()` function transforms data in all kinds of different formats into a standardized date format displaying year, then month, then day.

```{r}
# all of the different date formats that you can use with ymd

# "06 02 04"
# "06/02/04"
# "20060204"
#" 2006 2 4"
# 060204

ymd("2026-4-14")
ymd(20260414)
ymd("2026/4/14")
```

`mdy()` (month day year) and `dmy()` (day month year) formats also exist for dates.

```{r}
# an example with mdy()
mdy(04142026)


# an example with dmy()

dmy(14042026)

```

`ymd_hms()` helps to format datetimes.

```{r}

# ymd_hms() with "2020-04-01 10:30:13"

ymd_hms("2020-04-01 10:30:13")

# now with "2020/04/01 10.30"
ymd_hm("2020/04/01 10.30")

```

# Extracting Components

Once you have a date object, you can easily extract parts of it:

```{r}
birthday <- ymd("1983-04-14")

class(birthday)

# extract year using year()

year(birthday)


# extract month using month()
month(birthday)


# set label=TRUE to get the month name instead of number
month(birthday, label=TRUE)


# extract day using day()

day(birthday)


# want to know what day of the week??? use wday() with label=TRUE

wday(birthday, label=TRUE)


```

# Doing Math with Dates

Set up the object `today` as today's date:

```{r}
today <- Sys.Date()
```

Calculate the age based on today's date and your birthday. This is possible when both are in date or datetime formats. You can also use the function `difftime()` if you want to specify the units.

```{r}
#calc time beween today and birthday using simple subtraction

today-birthday

# calc time between birthday and today using difftime()

difftime(today, birthday, units="mins")
difftime(today, birthday, units="hours")

```

What will be the date 28 days from now?

```{r}

# add 28 to today

today+28
```

# Portal Data

The Portal Project is a long-term ecological study being conducted near Portal, AZ. Since 1977, the site has been used to study the interactions among rodents, ants and plants and their respective responses to climate.

```{r}
#LOAD DATA
portal_rodent <- read.csv("https://github.com/weecology/PortalData/raw/main/Rodents/Portal_rodent.csv")
```

Unfortunately, because the information about datetime is divided up into different columns, R does not recognize it as date/time data. What we need to do is combine and convert all of these columns into datetime. To do this, we can use the function `make_datetime()`

```{r}

# take a look at the data and how date/time are in the data

# use make_date() to turn these into a new column with datetime

portal_rodent_mod <- portal_rodent |>
  mutate(date=make_date(year, month, day))



```

A line plot showing the number of rodents captured per day over time:

```{r}

#use count(date) to tally up the number of rodents recorded on each day

portal_rodent_mod |>
  count(date) |>
  ggplot() +
  geom_line(aes(x=date, y=n)) +
  theme_minimal()

 
```

Use the `floor_date` function to round each date down to the first day of the month, which is great for time series grouping.

```{r}
# for floor date you need the date column and the unit which can be week, month, year, etc
# your unit needs to be in quotes (e.g. "month")

portal_rodent_mod |>
  mutate(month_floor = floor_date(date, "month")) |>
  count(month_floor) |>
  ggplot() +
  geom_line(aes(x=month_floor, y=n)) +
  theme_minimal()



```

Alternatively, use the `floor_date` function to round each date down to first day of the year:

```{r}

#same plot as above, but for "year"

portal_rodent_mod |>
  mutate(year_floor = floor_date(date, "year")) |>
  count(year_floor) |>
  ggplot() +
  geom_line(aes(x=year_floor, y=n)) +
  theme_minimal()


```

## In groups

Recreate the above daily line plot for 2017 only.

```{r}

portal_rodent_mod |>
  filter(year==2017) |>
  count(date) |>
  ggplot() +
  geom_line(aes(x=date, y=n)) +
  theme_minimal()


```

Recreate the above line plot March 15 - Dec 31, 2017.

```{r}

portal_rodent_mod |>
  filter(date > mdy(03152017),
         date < mdy(12312017)) |>
  count(date) |>
  ggplot() +
  geom_line(aes(x=date, y=n)) +
  theme_minimal()

portal_rodent_mod |>
  filter(date > "2017-03-15",
         date < "2017-12-31") |>
  count(date) |>
  ggplot() +
  geom_line(aes(x=date, y=n)) +
  theme_minimal()
```

Determine on which days of the week most rodents are observed (when most sampling happens). *Don't overthink it.*

```{r}
portal_rodent_mod |>
  mutate(wday=wday(date, label=TRUE)) |>
  count(wday)

```

Now turn your day of the week count into a barplot.

```{r}

portal_rodent_mod |>
  mutate(wday=wday(date, label=TRUE)) |>
  count(wday) |>
  ggplot() +
  geom_col(aes(x=wday, y=n, fill=wday)) +
  theme_minimal()

```

## Plotly

There really isn't much to this. Plotly provides a wrapper for existing ggplots.

First save the two above plots that you made in groups as an object.

```{r}

library(plotly)

#line_graph 

linegraph <- portal_rodent_mod |>
  mutate(month_floor = floor_date(date, "month")) |>
  count(month_floor) |>
  ggplot() +
  geom_line(aes(x=month_floor, y=n)) +
  theme_minimal()

#barplot 

bar_plot <- portal_rodent_mod |>
  mutate(wday=wday(date, label=TRUE)) |>
  count(wday) |>
  ggplot() +
  geom_col(aes(x=wday, y=n, fill=wday)) +
  theme_minimal()

```

Now let's translate them to plotly plots.

```{r}

# use ggplotly(objectname)

ggplotly(bar_plot)

ggplotly(linegraph)

```

Let's do something a little more complex now. Let's plot each year as a separate line on the line plot and see what plotly does with that.

```{r}
# need to use yday(date) to get day of year in order to plot separate years

plot1 <- portal_rodent_mod |>
  mutate(yday = yday(date)) |>
  group_by(year) |>
  count(yday) |>
  ggplot() +
  geom_point(aes(x=yday, y=n, color=year)) +
  theme_minimal()

ggplotly(plot1)

  
plot2 <- portal_rodent_mod |>
  mutate(yday = yday(date)) |>
  group_by(year) |>
  count(yday) |>
  ggplot() +
  geom_point(aes(x=yday, y=n, color=factor(year))) +
  theme_minimal()


```

Now let's play around with some bells and whistles. First let's add a range slider.

```{r}

# the code to add a slider: layout(xaxis = list(rangeslider = list(visible = TRUE)))

ggplotly(plot1) |>
  layout(xaxis = list(rangeslider = list(visible = TRUE)))


ggplotly(plot2) |>
  layout(xaxis = list(rangeslider = list(visible = TRUE)))

```

Let's try it with faceting.

```{r}
# first filter for years > 2010
# mutate a doy using yday()
# then make the counts


plot3 <- portal_rodent_mod |>
  filter(year > 2010) |>
  mutate(yday = yday(date)) |>
  group_by(year) |>
  count(yday) |>
  ggplot() +
  geom_point(aes(x=yday, y=n, color=factor(year))) +
  facet_wrap(~year) +
  theme_minimal()

ggplotly(plot3)

```

In order to make these work you can just render your .qmd to .html as normal and they will just work. If you want to export them as a separate thing, that's a lot more complicated.
