---
title: 'Aggregating data'
author: 
  - 'Casey Schine' 
  - 'adapted from Emily Malcolm-White'
format:
  html: 
    toc: TRUE
    code-overflow: wrap
    embed-resources: true
    code-tools:
      source: true
      toggle: false
      caption: none
execute: 
  message: FALSE
  warning: FALSE
---

```{r}
#LOAD PACKAGES 
library(tidyverse)
```

# `palmerpenguins` dataset

Size measurements, clutch observations, and blood isotope ratios for adult foraging Adélie, Chinstrap, and Gentoo penguins observed on islands in the Palmer Archipelago near Palmer Station, Antarctica. Data were collected and made available by Dr. Kristen Gorman and the Palmer Station Long Term Ecological Research (LTER) Program.

```{r}
#LOAD DATA 
library(palmerpenguins) 
data(penguins)         
```

## Remove rows with missing data with drop_na()

```{r}
# save as new object
# without pipe
penguins_mod <- drop_na(penguins)


# with pipe
penguins_mod <- penguins |>
  drop_na()

penguins_mod <- penguins |>
  filter(!is.na(bill_length_mm))

penguins_mod <- penguins |>
  filter(!is.na(sex))
```

::: callout-warning
Is it appropriate to remove rows with missing data? How many rows have missing data? Do the missing rows have something in common?

Removing rows can affect the validity and generalizability of your analysis!
:::

```{r}
# how can we tell what was in the rows that were removed?
penguins_mod <- penguins |>
  filter(!is.na(bill_length_mm))

penguins_mod <- penguins |>
  filter(!is.na(sex))

```


# `count()`

`count()` lets you quickly count the unique values of one or more variables. Suppose you want the number of penguins on each island.

```{r}
# use only count for island

penguins |>
  count(island)

# use count for island, species

penguins |>
  count(island, species)


# use count for island, species, sex
penguins |>
  count(island, species, sex)

# has weird NA groupings because sex is sometimes NA
# get a clearer picture after dropping NAs

penguins |>
  filter(!is.na(sex)) |>
  count(island, species, sex)

# you can drop the NAs before or after the count
# in some cases (not this one), that could give you different results

penguins |>
  count(island, species, sex) |>
  filter(!is.na(sex))

```

# `summarize()` or `summarise()` (either works)

Suppose we are interested in the average bill length of all Adelie penguins:

```{r}
# use summarize for average bill length in Adelie penguins
# Do it first without removing NAs

penguins |>
  filter(species=="Adelie") |>
  summarize(mean(bill_length_mm))
# have to deal with the NAs

# Then remove NAs
penguins |>
  filter(species=="Adelie") |>
  summarize(mean(bill_length_mm, na.rm=TRUE))

```

Suppose we are interested in the average bill length AND average bill depth of all Adelie penguins:

```{r}
# calculate average bill length and average bill depth for Adelie
# filter for Adelie then summarize

penguins |>
  filter(species=="Adelie") |>
  summarize(mean(bill_length_mm, na.rm=TRUE),
            mean(bill_depth_mm, na.rm=TRUE))

```

There are lots of other functions available:

-   `min`: minimum value
-   `max`: maximum value
-   `mean`: average or mean value
-   `median`: median value
-   `var`: variance
-   `sd`: standard deviation
-   `n`: count or number of values
-   `n_distinct`: counts number of distinct values

Suppose we are interested in the average bill length AND the median bill length of all Adelie penguins:

```{r}

# filter for Adelie and summarise (but change to median)
# average bill length and MEDIAN bill length

penguins |>
  filter(species=="Adelie") |>
  summarize(mean(bill_length_mm, na.rm=TRUE),
            median(bill_length_mm, na.rm=TRUE))

```

# `group_by()`

Let's say we were interested in the average bill length and bill depth of all penguin species in this dataset. We could repeat this for the other species (Gentoo and Chinstrap). This would be a fair amount of work AND the results would not end up in the same table.

OR we could use the `group_by` command!

```{r}
# group_by species and then summarize bill length and bill depth (mean)
penguins |>
  group_by(species) |>
  summarize(mean(bill_length_mm, na.rm=TRUE),
            mean(bill_depth_mm, na.rm=TRUE))

```

## Multiple Groups

Suppose we wish to have the average bill length and average bill depth broken down by sex AND species:

```{r}

# group by species/sex and then summarize avg bill depth
penguins |>
  group_by(species, sex) |>
  summarize(mean(bill_length_mm, na.rm=TRUE),
            mean(bill_depth_mm, na.rm=TRUE))

# do it first without filtering out NAs 

# do it again with NAs removed
```

# Recall: `mutate()`

For example, let's calculate create a new column which displays the body length weight in pounds (lbs) instead of grams. Recall: to convert from grams to pounds we need to multiply by 0.00220462

```{r}
# calculate body mass in lb from conversion factor, create new column

# then assign it to new object

penguins |>
  mutate(test_col = 1)

penguins_mod <- penguins |>
  mutate(test_col = body_mass_g*0.00220462)

```

# `case_when()`

Case when can be used in combination with `mutate` to create a new column with a categorical variable conditional on the values in another column.

For example:

```{r}
# penguins <- penguins |> 
#   mutate(penguin_length_cat = case_when(bill_length_mm > 50 ~ 'whoa! huge bill!', TRUE ~ '--' ))

penguins_mod <- penguins |>
  mutate(spp_abbr = case_when(species=="Adelie" ~ "A",
                               species=="Gentoo" ~ "G",
                               species=="Chinstrap" ~ "C"))

penguins_mod <- penguins |>
  mutate(big_penguin = case_when(body_mass_g > 5000 ~ "BIG",
                                 body_mass_g <= 5000 ~ "little"))

```

# Penguin Data Story

I love to play around with data and find little stories, which is what I will be asking you to do more and more as we move through the course material. So I had a little fun with the `palmerpenguins` data and this is what I came up with. 

Now let's look at the `penguins_raw` version of the data. There are a bunch more variables and below is an example of what you will be able to do by the end of next week.
```{r}
penguins_raw |>
  ggplot() +
  geom_point(aes(x=`Delta 13 C (o/oo)`, y=`Delta 15 N (o/oo)`, color=Species)) +
  theme_minimal()
```
This first plot shows the 15N isotopic ratio on the y-axis plotted against the 13C isotopic ratio on the x-axis. This first plot shows the points colored by species. There is a clear positive relationship between the 13C and 15N isotopic ratios with high 13C values corresponding with high 15N values. There is also a very clear grouping of each species along this gradient with Gentoo penguins having the lowest 13C and 15N values and Adelie penguins the highest. 

```{r}
penguins_raw |>
  ggplot() +
  geom_point(aes(x=`Delta 13 C (o/oo)`, y=`Delta 15 N (o/oo)`, color=`Body Mass (g)`)) +
  theme_minimal()
```
This second plot has the same x and y axes as the first plot, but now the points are colored by body mass. There is a clear relationship between body mass and the isotopic ratios of 13C and 15N with the highest body masses showing the lowest 15N and 13C ratios. Given the overall pattern my guess is that this relationship between body mass and isotopic signature is driven by the body mass of different species and species-specific differences in isotopic ratios rather than body mass having a direct relationship with isotopic ratio. 

If we then look at the distribution of body mass for each species:
```{r}
penguins_raw |>
  ggplot(aes(y=Species, x=`Body Mass (g)`)) +
  geom_boxplot() +
  geom_jitter(aes(color=Sex)) +
  theme_minimal()
```
We can see that Gentoo penguins have both the highest body mass and the lowest isotopic ratios. This relationship is also seen in the previous 2 figures. I added something on top of this boxplot called a jitter which is where all of the points that went into the boxplot are also plotted, which enables you to add a third variable. In this case I added sex, which allows us to see that within each species of penguin we have a clear difference in body mass between male and female penguins. 

This difference in isotopic ratios between penguin species is widely known and reflects the differences in diet. Gentoo penguins inhabit a lower trophic level and eat primarily krill (as reflected in their blood isotopes), while Adelie and Chinstrap penguins eat fish and inhabit a higher trophic level. If you want to get into the details about who trophic level is related to staple isotope ratios I am happy to chat!

What questions do these figures raise for you? What would you want to look at next?


