---
title:  'Making plots  with `ggplot2`: Barplots and Scatterplots'
author: 'Casey Schine adapted from Emily Malcolm-White'
format:
  html: 
    toc: TRUE
    code-overflow: wrap
    embed-resources: true
execute: 
  message: FALSE
  warning: FALSE
---

`ggplot2` is a package built within the `tidyverse` package for creating awesome graphs!

```{r}
library(tidyverse)
```

```{r}
#Import the can_lang dataset 
can_lang <- read.csv("https://raw.githubusercontent.com/ttimbers/canlang/master/inst/extdata/can_lang.csv")
```

# Recall our Top 10 example:

This code gave a list of 10 Aboriginal Languages which have the most number of people who speak them as their mother tongue:

```{r}

ten_lang <- can_lang |> 
  filter(category == "Aboriginal languages") |> 
  arrange(desc(mother_tongue)) |>
  select(language, mother_tongue) |> 
  slice(1:10)

```

# Barplots

Suppose we wanted to display this information in a barplot instead of in a table.

```{r}
#use geom_col
ggplot(data=ten_lang) +
  geom_col(aes(x=language, y=mother_tongue))


#send in data without using new object
can_lang |> 
  filter(category == "Aboriginal languages") |> 
  arrange(desc(mother_tongue)) |>
  #select(language, mother_tongue) |> 
  slice(1:10)|>
 ggplot() +
  geom_col(aes(x=language, y=mother_tongue))
```

Is there any improvements we could make to this graph?

# To better view text

Display the bars horizontally instead of vertically!

```{r}
# switch axes to change orientation
ggplot(data=ten_lang) +
  geom_col(aes(y=language, x=mother_tongue))

```

This flips the x and y axes!

# Labels, Colors, and Themes

```{r}
# add axis labels and plot title
ggplot(data=ten_lang) +
  geom_col(aes(y=language, x=mother_tongue)) +
  labs(x="Mother Tongue",
       y="Language")

ggplot(data=ten_lang) +
  geom_col(aes(y=language, x=mother_tongue)) +
  xlab("Mother Tongue") +
  ylab("Language")

# change bar colors
ggplot(data=ten_lang) +
  geom_col(aes(y=language, x=mother_tongue),
           color="darkolivegreen2",
           fill="purple") +
  labs(x="Mother Tongue",
       y="Language")

ggplot(data=ten_lang) +
  geom_col(aes(y=language, x=mother_tongue, fill=language)) +
  labs(x="Mother Tongue",
       y="Language")

# play around with themes
ggplot(data=ten_lang) +
  geom_col(aes(y=language, x=mother_tongue, fill=language)) +
  labs(x="Mother Tongue",
       y="Language") +
  theme_light()

# add a title
ggplot(data=ten_lang) +
  geom_col(aes(y=language, x=mother_tongue, fill=language)) +
  labs(x="Mother Tongue",
       y="Language") +
  ggtitle("This is a very nice plot") +
  theme_light()
  
```

::: callout-tip
Barplots are good for displaying one categorical variable and one numeric variable. The number variable could be counts (as above) or they could be averages or totals or maximums or minimums (or many other things!)
:::

# ggplot: scatterplot with `geom_point`

::: callout-tip
Scatterplots are good for displaying the relationship between two numerical variables.
:::

The `mtcars` dataset was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models). It's available inside the `ggplot` package which is already installed.

```{r}
#load the data
data(mtcars)

#what is in the dataset???
#?mtcars

```

## Color, Shape, and Size

```{r}
# make some plots
mtcars |>
ggplot() +
  geom_point(aes(x=wt, y=mpg))

# what should we plot?

# change shape using pch
#?pch

mtcars |>
ggplot() +
  geom_point(aes(x=wt, y=mpg),
             shape=15) +
  theme_light()

# change color outside of aes()

mtcars |>
ggplot() +
  geom_point(aes(x=wt, y=mpg),
             shape=15,
             color="#cb18db") +
  theme_light()

```

Note that you can change the color, shape (`pch` for plotting character) and size of these points!

### Quick update to the dataset

```{r}
#code to update `mtcars` dataset so that `am` is treated as a factor rather than a continuous numeric variable
mtcars <- mtcars |>  
  mutate(am_factor = as.factor(am)) 
```

This modifies the am column, which represents the transmission type of the car (0 = automatic, 1 = manual). The as.factor(am) function converts the am variable from a numeric type (0 or 1) into a categorical factor.

```{r}
# what is a factor and how does it change the plotting?

#plot using normal am
mtcars |>
ggplot() +
  geom_point(aes(x=wt, y=am),
             shape=15,
             color="#cb18db") +
  theme_light()

#plot using am_factor

mtcars |>
ggplot() +
  geom_point(aes(x=wt, y=mpg, color=am),
             shape=15) +
  theme_light()

mtcars |>
ggplot() +
  geom_point(aes(x=wt, y=mpg, color=am_factor),
             shape=15) +
  theme_light()

mtcars |>
ggplot() +
  geom_point(aes(x=wt, y=mpg, color=factor(am)),
             shape=15) +
  theme_light()

```

# Inside `aes()` or outside `aes()`?

What is the difference between these two graphs?

```{r}
ggplot(mtcars, aes(x=wt, y=mpg)) +
  geom_point(color="red") +
  theme_minimal()
```

```{r}
# global variables
ggplot(mtcars, aes(x=wt, y=mpg)) +
  geom_point() +
  geom_line() +
  theme_minimal()

#local variables
ggplot() +
  geom_point(data=mtcars, aes(x=wt, y=mpg)) +
  geom_line() +
  theme_minimal()
```

::: callout-tip
-   If the thing you are trying to change (color, shape, size, etc.) depends on a variable, you should put in *inside* the aesthetics
-   If the thing you are trying to change (color, shape, size, etc.) should happen for all things, you should not put it inside the aesthetics.
:::

# Customizing Colors in Aesthetics

We are going to take a look at three different ways to control colors for both discrete and continuous variables. First we are going to take a look at how to define colors for a discrete variable using the `am_factor` variable that we created above which takes on the value of either 0 or 1.

```{r}
# color in aesthetics

# for discrete variables
# ggplot(mtcars, aes(x=wt, y=mpg)) +
#   geom_point(aes(color=am_factor)) +
#   scale_color_manual(values=c("black", "orange")) 

mtcars |>
ggplot() +
  geom_point(aes(x=wt, y=mpg, color=factor(am)),
             shape=15) +
  scale_color_manual(values=c("black", "orange")) +
  theme_light()

mtcars |>
  mutate(am_words = case_when(am==0 ~ "no",
                              am==1 ~ "yes")) |>
ggplot() +
  geom_point(aes(x=wt, y=mpg, color=am_words),
             shape=15) +
  scale_color_manual(values=c("black", "orange")) +
  theme_light()

```

Now let's take a look at two different ways to do this for a discrete variable. First with the `scale_color_gradient()` function that you should already have access to with the packages that you have loaded.

```{r}
# for continuous variables
# without loading in new library
# scale_color_gradient(high=, low=)
# scale_color_gradient(low  = "lightblue",high = "darkblue")
# what colors are available to you? How do you get them/name them?

mtcars |>
ggplot() +
  geom_point(aes(x=wt, y=mpg, color=qsec),
             shape=15) +
  scale_color_gradient(high="purple", low="linen") +
  theme_light()

```

There are lots of ways to control the colors and palette and we will get into how to operate more of those controls in a few classes.

# Global vs. Local Aesthetics

Global aesthetic mappings apply to all geometries and can be defined when you initially call `ggplot()`. All the geometries added as layers will default to this mapping. Local aesthetic mappings add additional information or override the default mappings.

```{r}
#color = am as a global aesthetic

mtcars |>
ggplot(aes(x=wt, y=mpg, color=qsec)) +
  geom_point() +
  scale_color_gradient(high="purple", low="linen") +
  theme_light()

```

```{r}
#color = am as a local aesthetic
mtcars |>
ggplot() +
  geom_point(aes(x=wt, y=mpg, color=qsec)) +
  scale_color_gradient(high="purple", low="linen") +
  theme_light()
```

```{r}
# overwriting color = am as a global aesthetic with a local aesthetic
mtcars |>
ggplot(aes(x=wt, y=mpg, color=qsec)) +
  geom_point(color="magenta") +
  scale_color_gradient(high="purple", low="linen") +
  theme_light()

```

# In-Class Excercise:

## Plot 1

Create a scatterplot using `geom_point()` to explore the relationship between engine displacement (`disp`) and horsepower (`hp`) in the `mtcars` dataset. In addition to plotting these two numeric variables, *map a third numeric variable* of your choice to color. Apply a non-default continuous color scale (either `scale_color_gradient()` or `scale_color_viridis_c()`), and include a clear title, axis labels, and a non-default theme. After creating your plot, write 2–3 sentences describing the pattern you observe between displacement and horsepower.

```{r}

```

## Plot 2

Now extend the story by investigating a different relationship in the dataset. Choose two numeric variables other than `hp` and `disp` that you think might help explain fuel efficiency or vehicle performance, and create a new scatterplot using `geom_point()`. Convert one relevant grouping variable (such as `am`, `cyl`, or `gear`) to a factor and map it to color. Apply a non-default discrete color scale using `scale_color_manual()`, and include a clear title, axis labels, and a theme. In 2–3 sentences, explain why you chose these variables and what you observe in the plot.

```{r}

```
