---
title: "Midterm Prep"
subtitle: 'Thursday: How to turn exploration into a clear analysis'
author: "Casey Schine"
format:
  html: 
    toc: TRUE
    code-overflow: wrap
    embed-resources: true
    code-tools:
      source: true
      toggle: false
      caption: none
execute: 
  message: FALSE
  warning: FALSE
editor_options: 
  chunk_output_type: console
---

# 1. Recap

### Workflow:

Question → Plot → Pattern → Interpretation

```{r}
library(tidyverse)
library(scales)
library(ggthemes)

#code to read in data
getwd()
# EJI
library(sf)
eji_data <- st_read("~/Dropbox/!Middlebury/Teaching/STAT 118/S2026/S2026_R_Proj/midterm/eji_sf.gpkg")

# YRBS
yrbs_data <- read_csv("~/Dropbox/!Middlebury/Teaching/STAT 118/S2026/S2026_R_Proj/midterm/yrbss_2024.csv")
```

Plot we made using the EJI dataset looking at the relationship between smoke exposure and cancer levels.

```{r}
#variables to explore smoke v. cancer
#E_SMOKE
#E_CANCER
eji_data |>
  #filter(STATE_NAME=="California") |>
  ggplot() +
  geom_point(aes(x=E_SMOKE, y=E_CANCER), alpha=0.05) +
  ggtitle("Canver vs. Wildfire Smoke Scatterplot with all data") +
  #scale_x_log10() +
  labs(x="Avg. Annualized Frequency of Smoky Days",
       y="Individuals with Cancer (%)") +
  theme_tufte()

eji_data |>
  filter(E_SMOKE>15) |>
  ggplot(aes(x=E_SMOKE, y=E_CANCER)) +
  geom_hex(bins=50) +
  geom_smooth(color="red") +
  scale_fill_viridis_c() +
   ggtitle("Canver vs. Smoke 2D density plot with smoke threshold at 15") +
  scale_x_log10() +
  theme_tufte()

eji_data |>
  #filter(E_SMOKE>15) |>
  ggplot(aes(x=E_BURN, y=E_CANCER)) +
  geom_hex(bins=50) +
  geom_smooth(color="red") +
  scale_fill_viridis_c() +
  ggtitle("Canver vs. Burn Area 2D density plot with all data") +
  scale_x_log10() +
  theme_tufte()

# Question: What does California look like compared with the entire population?
eji_data |>
  filter(STATE_NAME=="California") |>
  ggplot() +
  geom_point(data=eji_data, aes(x=E_SMOKE, y=E_CANCER), alpha=0.05) +
  ggtitle("Canver vs. Wildfire Smoke Scatterplot with all data, California in Red") +
  #scale_x_log10() +
  labs(x="Avg. Annualized Frequency of Smoky Days",
       y="Individuals with Cancer (%)") +
  geom_point(aes(x=E_SMOKE, y=E_CANCER), alpha=0.2, color="red") +
  theme_tufte()

# Same as last plot but histogram 
eji_data |>
  mutate(is_cali = case_when(STATE_NAME=="California" ~ "CA",
                             STATE_NAME!="California" ~ "Not CA")) |>
  ggplot() +
  geom_histogram(aes(x=E_SMOKE, fill=is_cali), alpha=0.5, position="identity") +
  #scale_x_log10() +
  ggtitle("Wildfire Smoke Histogram with all data, California in Red") +
  labs(x="Avg. Annualized Frequency of Smoky Days") +
  theme_tufte()

# Same as last plot but density plot to deal with very different quantities of data
eji_data |>
  mutate(is_cali = case_when(STATE_NAME=="California" ~ "CA",
                             STATE_NAME!="California" ~ "Not CA")) |>
  ggplot() +
  geom_density(aes(x=E_SMOKE, fill=is_cali, color=is_cali), alpha=0.5, position="identity") +
  #scale_x_log10() +
  ggtitle("Wildfire Smoke Histogram with all data, California in Red") +
  labs(x="Avg. Annualized Frequency of Smoky Days") +
  theme_tufte()

# still working with smoke but now histogram of all states with facet
eji_data |>
  ggplot() +
  geom_histogram(aes(x=E_SMOKE, fill=STATE_NAME), alpha=0.5, position="identity") +
  #scale_x_log10() +
  ggtitle("Wildfire Smoke Histogram by STATE") +
  labs(x="Avg. Annualized Frequency of Smoky Days") +
  facet_wrap(~STATE_NAME) +
  theme_tufte() +
  theme(legend.position="none")
  
```

### In your report:

Introduction → dataset + question

Results → plots + patterns

Discussion → interpretation

# 2. Group check-in: What questions are you exploring?

Each person briefly share, what question you are currently exploring and how you are approaching it. Help your group mates clarify and refine their questions.

# 3. Question → Plot

Plot choice is part of the reasoning process, not simply decoration. You should be picturing the plot that you want to make and then matching it with the skills needed o make it. Rather than the other way around. Drawing it out on paper and breaking down the steps you will need to make it can be very helpful.

Give me some example questions.

What kind of plot would help answer this?

# 4. Turning messy plots into clear plots

Let's take a rough exploratory plot and improve it.

Possible improvements:

-   reorder categories
-   simplify colors
-   improve axis labels
-   add clear titles
-   remove unnecessary clutter

Explain:

Your first plots help you explore. The plots in your report should help the reader understand the pattern.

# 5. Results versus Discussion

What do you see in this plot?

```{r}
#make example plot
```

Write down answers and separate into what belongs in the results and what belongs in the discussion.

### Results = describe pattern

### Discussion = explain pattern

# 6. Studio work

Work on your project with your group. Ask questions and help each other debug code.

# 7. End-of-class checkpoint

Finish these statements:

-   My question is...
-   My two key plots will show...
-   The main pattern I see is...
