---
title: "STAT 118: Homework"
subtitle: "Working with text with `stringr`"
author: "YOUR NAME HERE"
format:
  html: 
    toc: true
    code-overflow: wrap
    code-fold: true
    embed-resources: true
---

```{r}
#LOAD PACKAGES 
library(tidyverse)

#LOAD DATA
courses <- read_csv("~/Dropbox/!Middlebury/Teaching/STAT 118/S2026/S2026_R_Proj/Homework/courses_S26-2.csv")
courses <- courses |>
  mutate(isCW = str_extract(distros, "CW")) %>%
  mutate(distros = str_remove(distros, "CW"))
```

# 1.
Create new column called `isDED` which indicates courses which have "DED" listed inside the `distros` column. 

```{r}
courses <- courses |>
  mutate(isDED = (str_detect(distros, "DED")))


```


# 2.

Create a new column(s) called `courseDEPT` and `courseCODE` and `courseSECT` which seperate the the department course identifier (ie. STAT0118B-S26) into three separate pieces (ie. STAT, 0118, and A). 

```{r}
courses <- courses |>
  mutate(courseDEPT = (str_sub(course_code, start=1, end=4))) |>
  mutate(courseCODE = (str_sub(course_code, start=5, end=8))) |>
  mutate(courseSECT = (str_sub(course_code, start=9, end=9)))

```

# 3.

How many classes will be offered in Spring 2026 which satisfy the DED requirements? (Classes with multiple sections will contribute several classes to the count).

```{r}
courses |>
  count(isDED)

```


# 4.
How many *unique* classes will be offered in Spring 2026 which satisfy the DED requirements? (Classes with multiple sections will only count as one class).

```{r}
test <- courses |>
  filter(isDED) |>
  distinct(courseDEPT, courseCODE) |>
  count(courseDEPT)

test2 <- courses |>
filter(courseSECT=="A") |>
  filter(isDED) |>
  filter(courseDEPT=="SOCI")

```


# 5.
Suppose you are interested in taking a class which satisfies one of the cultures and civilization requirement (SOA, NOA, MDE, SAF, EUR, or AMR). Which classes will be offered in Spring 2026 which satisfy at least one cultures and civilization tag?

```{r}

courses|>
  filter(str_detect(distros, "SOA|NOA|MDE|SAF|EUR|AMR"))

```


# 6.
A "unicorn class" is defined as a course which is able to satisfy three different distribution requirements -- a college writing class, an academic distribution requirement, and a regional requirement. Create a list of all the unicorn classes offered in Spring 2026.

```{r}
courses |>
  filter(isCW == "CW") |>
  filter(str_detect(distros, "SOA|NOA|MDE|SAF|EUR|AMR")) |>
  filter(str_detect(distros, "DED|HIS|LIT|SCI|SOC|ART|PHL|LNG"))

```

# 7.
Suppose you are interested in taking a class that meets on Wednesday nights (and only Wednesday nights). Are there any classes offered in Spring 2026 that meet this criteria? 
```{r}
courses |>
  mutate(start_hour = str_sub(time, start = 1, end = 7)) |>
  filter(str_detect(time, "Wednesday"), 
         !str_detect(time, "Monday|Tuesday|Thursday|Friday"), 
         str_detect(start_hour, "pm"),
         str_detect(start_hour, "6:|7:|8:|9:|10:|11:")) |>
  pull(title)

```

# 8.
Suppose we are interested in how many classes offer 1 section, how many offer 2 sections, how many offer 3 sections, etc. First, calculate the number of classes which offer 1 sections, 2 sections, 3 sections, etc. 

```{r}
section_courses <- courses |>
  group_by(courseDEPT, courseCODE) |>
  mutate(n_sections = n())

section_courses |>
  group_by(n_sections) |>
  summarize(n_classes = n())

```


# 9.
Which class(es) offer(s) the most number of sections? 

```{r}
section_courses |>
  filter(n_sections == 5)

```
