---
title: "STAT 118: Homework C"
subtitle: "Aggregating data"
author: "Casey Schine"
format:
  html:
    toc: true
    code-overflow: wrap
    code-fold: true
    embed-resources: true
execute: 
  warning: false
  message: false
---

```{r }
#LOAD PACKAGES 
library(tidyverse)
```

```{r}
#LOAD DATASET
# This iris dataset should be built into base R (no need to download a package). Just run the code below to load it. 
data("iris")
```

This famous (Fisher's or Anderson's) iris data set gives the measurements in centimeters of the variables sepal length and width and petal length and width, respectively, for 50 flowers from each of 3 species of iris. The species are Iris setosa, versicolor, and virginica.

# 1. 

Verify using `group_by` and `summarize` that there are indeed 50 different flowers from each of the different species of iris. 
```{r}
iris |> 
  group_by(Species) |>
  summarize(n())
```



# 2. 

Create a table which displays the average sepal length and average sepal width for the setosa species of iris. No other data should be in this table. 

```{r}
iris |>
  filter(Species=="setosa") |>
  summarize(average_sepal_length=mean(Sepal.Length),
            average_sepal_width=mean(Sepal.Width))
```


# 3. 

Create a table which displays the average sepal length and the median sepal length for the setosa species of iris. No other data should be in this table. 
```{r}
iris |>
  filter(Species=="setosa") |>
  summarize(average_sepal_length=mean(Sepal.Length),
            median_sepal_length=median(Sepal.Length))
```




# 4. 

Create a table which displays the minimum sepal length and the maximum sepal length for each of the three different species of iris. 

```{r}
iris |>
  group_by(Species) |>
  summarize(min_sepal_length=min(Sepal.Length),
            max_sepal_length=max(Sepal.Length))
```




# 5. 

Create one table which displays the minimum sepal length and the maximum sepal length for the species setosa and versicolor. Note that this table should *not* include any information about the virginica species. (This might be a challenge, but you can do it!)

```{r}

iris |>
  filter(Species!= "virginica") |> #lots of different way to do this step
  group_by(Species) |>
  summarize(min_sepal_length=min(Sepal.Length),
            max_sepal_length=max(Sepal.Length))

```




