---
title: "STAT 118: Homework G"
subtitle: 'Maps with `maps` and `sf`'
author: "YOUR NAME HERE"
format:
  html: 
    toc: true
    code-overflow: wrap
    code-fold: true
    embed-resources: true
---

```{r message=FALSE, warning=FALSE}
#LOAD PACKAGES 
library(tidyverse)
library(sf) #this is a package needed so R can work with sf objects
```

```{r message=FALSE, warning=FALSE}
#LOAD DATA
library(spData) #this packages contains the dataset (with sf objects) that we will be using today
data("us_states")

library(rnaturalearth)
library(rnaturalearthdata)
world <- ne_countries(scale = "medium", returnclass = "sf")
africa <- world[world$continent == "Africa", ]

```

# 1.
Create a basic map of the countries of the world. To access `sf` geometries for the world use the `world` dataset loaded above. The land should be some shade of green. Use something other than the default ggplot theme. 

```{r}


```



# 2. 

Use the columns `pop_est` and `gdp_md` to create a new column called `gdp_per_cap` which contains the GDP per capita for each country in the world in this dataset. 

```{r}

```


# 3.
Create a choropleth map which colors each country depending on the GDP per capita. 

- Utilize an appropriate color palette for this data
- Add a theme
- make sure the legend labels are clear to a general audience (you might need to research this a bit...)
- the country borders should be "black"

```{r}

```

# 4.
Create a choropleth map which colors each country based on their economy type (`economy`). 

- Utilize an appropriate color palette for this data
- Add a theme
- make sure the legend labels are clear to a general audience (you might need to research this a bit...)
- the country borders should be "grey"

```{r}

```
`
# 5. 

Create a choropleth map of Africa based on the population in each country. 
Use `scale_fill_viridis_c(option = "viridis")`. 

```{r}

```



# 6. 

Create a choropleth map of Africa based on the population in each country. 
Use `scale_fill_viridis_c(option = "viridis", trans = "log")`. 

```{r}

```

# 7. 

Why is the trans = "log" transformation applied? 


# 8. 

Add country names to your map from #6. 

```{r}

```

# In-Class Exercise: Comparing Two County-Level Rates

*Copy your work from class here.*

Today you will create and compare two variables using the North Carolina dataset.

## Step 1: Create Two Proportions (1974)

Add the following variables to your dataset:

- Proportion of nonwhite births, calculated as `NWBIR74 / BIR74`
- SIDS deaths per birth, calculated as `SID74 / BIR74`

Why are we dividing by total births instead of mapping raw counts?

## Step 2: Map Each Variable

Create two separate choropleth maps: 1) Proportion of nonwhite births, 2) SIDS deaths per birth.

Use a continuous color scale using the techniques we just learned in class, and use anything other than the default ggplot theme. Add a title to your plot and name your colorbar.

```{r}

```


## Step 3: Compare the Maps

Write a short paragraph (3–4 sentences) answering the questions: 

- Do these variables show similar spatial patterns?
- Where do they look similar or different?
- What is one limitation of comparing county-level maps like this?


