---
title: "Homework: Practicing New Plot Types"
subtitle: 'with small multiples and patchwork'
author: "YOUR NAME HERE"
format:
  html: 
    toc: true
    code-overflow: wrap
    code-fold: true
    embed-resources: true
---

## Instructions

Use the `diamonds` dataset from `ggplot2`.

Follow the steps below exactly. The goal is to practice using new plotting tools.

For each section:
- Create the plot
- Make the required modifications
- Write 1–2 sentences describing what you see
- Use something other than the default theme. But don't be too cute, your plots still need to be legible. 

```{r}
library(tidyverse)
library(ggplot2)
library(ggridges)
library(patchwork)

data(diamonds)
```

## 1. Heatmap (Categorical × Categorical)

Create a heatmap showing the **proportion of clarity within each cut**. You will first need to create a variable for proportion using `mutate`.


```{r}
df_counts <- diamonds |>
  count(cut, clarity) |>
  group_by(cut) |>
  mutate(prop = n / sum(n))

diamonds |>
  count(cut, clarity) |>
  group_by(cut) |>
  mutate(prop = n / sum(n))

p1 <- ggplot(df_counts, aes(x = cut, y = clarity, fill = prop)) +
  geom_tile()

p1

diamonds |>
  group_by(cut, clarity) |>
  summarize(n=n()) |>
  mutate(prop=n/sum(n))
```

**Required modifications**
- Change the legend title to "Proportion"
- Add a plot title
- Reorder `clarity` so the levels appear in a meaningful order (unless you are very familiar with diamond clarity categories, you will need to do a little research to develop a meaningful order)

**Write (1–2 sentences)**
- What does the color represent?
- What pattern do you observe?

## 2. Distribution Across Groups (Ridgeline Plot)

Compare the distribution of **price across cut**.

```{r}
p2 <- ggplot(diamonds, aes(x = price, y = cut, fill = cut)) +
  geom_density_ridges()

p2
```

**Required modifications**
- Set `alpha` to less than 1
- Remove the legend
- Add a clear x-axis label

**Write (1–2 sentences)**
- How does price distribution differ across cuts?

## 3. 2D Density Plot

Visualize the relationship between **carat and price**.

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point()

p3 <- ggplot(diamonds, aes(x = carat, y = price)) +
  geom_density_2d_filled()
```

**Required modifications**
- Zoom to the main data region using `coord_cartesian()` (we have *not* discussed how to do this in class, you will need to figure out how on your own, the help documentation for the function is a great place to start).
- Add a plot title
- Add readable axis labels

**Write (1–2 sentences)**
- Where are most observations concentrated?
- What relationship do you see?

## 4. Combined Figure (Patchwork)

Combine your three plots into a multi-panel figure.

```{r}
(p1 | p2) / p3 +
  plot_layout(guides = "collect", heights = c(1, 2)) +
  plot_annotation(title = "Your title here")
```

**Requirements**
- Heatmap and ridgeline plot in the first row
- 2D density plot in the second row
- Second row taller than the first
- Legends collected
- Overall title added

**Write (1–2 sentences)**
- What does each panel show?
- What is the overall takeaway?

***Remember to give your rendered .html a quick look for formatting before submitting.***

