---
title: "Homework: Practicing New Plot Types"
subtitle: 'with small multiples and patchwork'
author: "YOUR NAME HERE"
format:
  html: 
    toc: true
    code-overflow: wrap
    code-fold: true
    embed-resources: true
---

## Instructions

Use the `diamonds` dataset from `ggplot2`.

Follow the steps below exactly. The goal is to practice using new plotting tools.

For each section:

- Create the plot
- Make the required modifications
- Write 1–2 sentences describing what you see
- Use something other than the default theme. But don't be too cute, your plots still need to be legible. 
- Manipulate the color palette on all plots to be something other than the default palette. 

```{r}
library(tidyverse)
library(ggplot2)
library(ggridges)
library(patchwork)

data(diamonds)
```

## 1. Heatmap (Categorical × Categorical)

Create a heatmap showing the **proportion of clarity within each cut**. You will first need to create a variable for proportion of each clarity class within each cut type using `mutate`.


```{r}

```

**Required modifications**

- Change the legend title to "Proportion" and shift the legend below the heatmap. 
- Add a plot title
- Reorder `clarity` so the levels appear in a meaningful order (unless you are very familiar with diamond clarity categories, you will need to do a little research to develop a meaningful order).

**Write (1–2 sentences)**

- What does the color represent?
- What pattern(s) do you observe?

## 2. Distribution Across Groups (Ridgeline Plot)

Compare the distribution of **price across cut**.

```{r}

```

**Required modifications**

- Set `alpha` to less than 1
- Remove the legend
- Add a clear x-axis label

**Write (1–2 sentences)**

- How does price distribution differ across cuts?

## 3. 2D Density Plot

Visualize the relationship between **carat and price**.

```{r}

```

**Required modifications**

- Zoom to the main data region using `coord_cartesian()` (we have *not* discussed how to do this in class, you will need to figure out how on your own, the help documentation for the function is a great place to start).
- Add a plot title
- Add readable axis labels

**Write (1–2 sentences)**

- Where are most observations concentrated?
- What relationship do you see?

## 4. Combined Figure (Patchwork)

Combine your three plots into a multi-panel figure.

```{r}

```

**Requirements**

- Heatmap and ridgeline plot in the first row
- 2D density plot in the second row
- Second row taller than the first
- Legends collected
- Overall title added

**Write (1–2 sentences)**

- What does each panel show?
- What is the overall takeaway?

***Remember to give your rendered .html a quick look for formatting before submitting.***

