---
title: "STAT 118: Homework I"
subtitle: 'Joining tables with dplyr'
author: "YOUR NAME HERE"
format:
  html: 
    toc: true
    code-overflow: wrap
    code-fold: true
    embed-resources: true
---

```{r message=FALSE, warning=FALSE}
#LOAD PACKAGES 
library(tidyverse)

#Load gapminder data
library(gapminder)
data(gapminder)
```

# 0.
Download the `co2.csv` file from Canvas. Import the dataset into this assignment in a dataframe called `co2`. 

```{r}
co2 <- read_csv("/Users/caseysmith/Dropbox/!Middlebury/Teaching/STAT 118/S2026/S2026_R_Proj/Homework/co2.csv")

```

# 1.
Use your brain and skills to figure out what the code below is doing. Add a sentence describing what this does. 

```{r}
country_continent <- gapminder %>% 
  select(country, continent) %>% 
  distinct()
```

# 2.
What columns do `country_continent` and `co2` have in common? (*This is your key*)

**country**

# 3.
What columns does `country_continent` have that `co2` doesn't? 

**continent**

# 4.
What columns does `co2` have that `country_continent` doesn't? 

**year and co2_emissions_pc**

# 5.
Create a new table called `co2_joined`, which includes all columns from `co2` in addition to the column `continent`. To start, use `inner_join()`. Make a note of how many rows you have in the new table, and compare it with the original tables.

```{r}

co2_joined <- co2 |>
  inner_join(country_continent)

co2_joined <- co2 |>
  inner_join(country_continent, by="country")

```

**original table has 22,892 obs and new table has 15930 obs**

# 6.
Create a new table called `co2_joined_left`, which includes all columns from `co2` in addition to the column `continent` but this time use `left_join()` instead of `inner_join()`. Make a note of how many rows you have in the new table, and compare it with the original tables.

```{r}

co2_joined_left <- co2 |>
  left_join(country_continent)

```

**original table has 22,892 obs and new table has 22,892 obs**


# 7. 
What columns do `gapminder` and `co2` have in common? 

**country and year**

# 8.
What columns does `gapminder` have that `co2` doesn't? 

**continent, lifeExp, pop, gdpPercap**

# 9.
What columns does `co2` have that `gapminder` doesn't? 

**co2_emissions_pc**

# 10. 
Create a new table called `gapminder_joined`, which includes only *complete* rows from data from both tables joined together by their common variable keys (use `inner_join()`). Make a note of how many rows you have in the new table, and compare it with the original tables.

```{r}

gapminder_joined <- co2 |>
  inner_join(gapminder, by=join_by(country, year))

```

**original co2 table has 22,892 obs, original gapminer has 1,704 obs, and new table has 1,620 obs**


# 11. 
Use the `anti_join()` function to identify which rows (if any) in the `gapminder` dataset do NOT have a match in out `co2` dataset. 

```{r}

gapminder |>
  anti_join(co2, by=join_by(country, year))


```

# 12. 
Use the `anti_join()` function to identify which rows (if any) in the `co2` dataset do NOT have a match in out `gapminder` dataset.

```{r}

co2 |>
  anti_join(gapminder, by=join_by(country, year))



```


# 13.
Use `gapminder_joined` to create a scatterplot of  `gdpPercap` vs. `co2_emissions_pc` for any one year. 

```{r}
library(plotly)

plot <- gapminder_joined |>
  filter(year==2002) |>
  ggplot(aes(x=gdpPercap, y=co2_emissions_pc, label=country)) +
  geom_point()

ggplotly(plot)

```



